---
title: "POL 212: Final Project"
---

```{r setup, include=FALSE}
library(knitr)
library(here)

opts_knit$set(root.dir = here::here("01_thesis_replication"))
opts_chunk$set(echo = TRUE)
```

```{r}
# Load necessary libraries
library(tidyverse)
library(dplyr)
library(labelled)
library(kableExtra)
library(sjlabelled)
library(stringr)
library(haven)
library(readxl)
library(stargazer)
library(broom)
library(gt)

# Problem with install yikes

# devtools::install_github("lapop-central/lapop", force = TRUE, build_vignettes = TRUE)
# library(lapop)

# Load the data
LAPOP <- haven::read_dta(here("01_data/02_lapop/SLV_2021_LAPOP.dta")) # Public Opinion data from 2021
violence = read_csv(here("01_data/ACLED Data_2025-11-20.csv"))

LAPOP_2019 = haven::read_dta(here("01_data/02_lapop/ElSalvador LAPOP AmericasBarometer 2018 v1.0_W.dta")) # Public Opinion data from 2019

# agrreggated / cleaned data
clean_merged = read_rds(here("01_data/01_juan_data/clean-merged-with-votes.rds"))

clean_merged = clean_merged %>%
  rename(municipio = name)

# Clean data

# Sort LAPOP into Group A and Group B

LAPOP_A = LAPOP %>%
  filter(core_a_core_b == "Core A") # Group A respondents


```

Logit with 2018/2019 Data

T - EDR

idio2 - Do you think that your economic situation is better than, the same as, or worse than it was 12 months ago?

l1 - Now, to change the subject... On this card there is a 1-10 scale that goes from left to right. The number one means left and 10 means right. Nowadays, when we speak of political leanings, we talk of those on the left and those on the right. In other words, some people sympathize more with the left and others with the right. According to the meaning that the terms "left" and "right" have for you, and thinking of your own political leanings, where would you place yourself on this scale? Tell me the number.

jc10 - Some people say that under some circumstances it would be justified for the military of this country to take power by a coup d'état (military coup). In your opinion would a military coup be justified under the following circumstances? When there is a lot of crime

jc13 - Some people say that under some circumstances it would be justified for the military of this country to take power by a coup d'état (military coup). In your opinion would a military coup be justified under the following circumstances? When there is a lot of corruption

jc15a - Do you believe that when the country is facing very difficult times it is justifiable for the president of the country to close the Legislative Assembly and govern without Legislative Assembly?

jc16a - Do you believe that when the country is facing very difficult times it is justifiable for the president of the country to dissolve the Supreme Court of Justice and govern without the Supreme Court of Justice?

vic1ext - Have you been a victim of any type of crime in the past 12 months? That is, have you been a victim of robbery, burglary, assault, fraud, blackmail, extortion, violent threats or any other type of crime in the past 12 months?

vic1exta - \^ how many times

vicbar7 - Have there been any murders in that last 12 months in your neighborhood?

vicbar7f - How many times did this occur?

fear6fa - For fear of being a victim of crime, in the last twelve months ... Have you kept your minor children from going to school for fear of their safety?

vicbar4a - Have you or anyone in your immediate family (children, spouse, wife) been a victim of extortion in the last 12 months?

aoj11 - Speaking of the neighborhood where you live and thinking of the possibility of being assaulted or robbed, do you feel very safe, somewhat safe, somewhat unsafe or very unsafe?

paz1 - Do you consider that the Peace Accords have been very good, good, bad, or very bad for the country?

ing4 - Changing the subject again, democracy may have problems, but it is better than any other form of government. To what extent do you agree or disagree with this statement?

aoj22new - To reduce crime in a country like ours, penalties for criminals need to increase. Do you agree or disagree with this statement?

dem30 - In your opinion, [country] is a democracy?

dem11 - Do you think that our country needs a government with an iron fist, or do you think that problems can be resolved with everyone's participation?

exc2 - Has a police officer asked you for a bribe in the last twelve months?

M1. Hablando en general acerca del gobierno actual, ¿diría usted que el trabajo que está realizando el Presidente Nayib Bukele es...?: [Leer alternativas] (1) Muy bueno (2) Bueno (3) Ni bueno, ni malo (regular) (4) Malo (5) Muy malo (pésimo) (888888) [NO LEER] No sabe (988888) [NO LEER] No responde (999999) [NO LEER] Inaplicable

T ING4. Puede que la democracia tenga problemas, pero es mejor que cualquier otra forma de gobierno. ¿Hasta qué punto está de acuerdo o en desacuerdo con esta frase? [SOLO LEER DE SER NECESARIO: Por favor, dígame un número entre 1 y 7]

Going to run a logit on M1 p much

```{r}

# make a positive or negative perception of bukele variable 
LAPOP = LAPOP %>%
  mutate(
    bukele_perception = case_when(
      m1 %in% c(1,2) ~ "positive",  # Positive perception
      m1 == 3 ~ "neutral",      # Neutral perception
      m1 %in% c(4,5) ~ "negative",  # Negative perception
    ),
    bukele_perception_dummy = case_when(
    m1 %in% c(1,2) ~ 1,  # Positive perception
    m1 %in% c(3,4,5) ~ 0,  # Negative or Neutral perception
    TRUE ~ NA_integer_    # Handle neutral and missing values
  ))

# Make a little histogram


LAPOP |> 
  filter(!is.na(bukele_perception)) |>
  ggplot(aes(x = bukele_perception)) +
    geom_bar(fill = "steelblue") +
    labs(title = "Perception of President Nayib Bukele",
         x = "Perception Category",
         y = "Count") +
    theme_minimal()

# count how many with negaitve vs positive 
sum(LAPOP$bukele_perception_dummy == 1, na.rm = TRUE)  # Positive perception
sum(LAPOP$bukele_perception_dummy == 0, na.rm = TRUE)  # Negative perception
sum(LAPOP$bukele_perception == 'neutral', na.rm = TRUE)  # Positive perception
```

```{r}
# Run logistic regression

# Basic Model: Violence Exposure (vic1ext), Urbanization, Economic Perspective (idio2) Is country a Democracy (dem30), Close legislature (jc15a)

sum(complete.cases(LAPOP[, c("bukele_perception_dummy", "vicbar4a")]))

logit_model <- glm(bukele_perception_dummy ~ vicbar4a + ur1new + idio2 + dem30,
            data = LAPOP,
            family = binomial(link = "logit")
)

stargazer::stargazer(logit_model, digits = 5, type = "text")

exp(coef(logit_model))


logit_tidy <- tidy(logit_model, conf.int = TRUE, exponentiate = TRUE) |>
  mutate(term = case_when(
    term == "(Intercept)" ~ "Intercept",
    term == "vicbar4a" ~ "Victimization Index",
    term == "ur1new" ~ "Crime Perception",
    term == "idio2" ~ "Ideology (Left–Right)",
    term == "dem30" ~ "Demographics Index",
    term == "jc15a" ~ "Trust in Institutions",
    TRUE ~ term
  ))

gt(logit_tidy) |>
  fmt_number(columns = c(estimate, conf.low, conf.high), decimals = 3) |>
  cols_label(
    term = "Variable",
    estimate = "Odds Ratio",
    std.error = "Std. Error",
    statistic = "z-value",
    p.value = "p-value",
    conf.low = "CI Lower",
    conf.high = "CI Upper"
  ) |>
  tab_header(
    title = "Logistic Regression Predicting Support for Bukele",
    subtitle = "Odds ratios with 95% confidence intervals"
  )


logit_plot <- tidy(logit_model, conf.int = TRUE, exponentiate = TRUE) |>
  filter(term != "(Intercept)") |>
  mutate(term = factor(term, levels = rev(unique(term))))

ggplot(logit_plot, aes(x = term, y = estimate)) +
  geom_point(size = 3, color = "steelblue") +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
  coord_flip() +
  labs(
    title = "Effects on Probability of Supporting Bukele",
    subtitle = "Odds ratios with 95% confidence intervals",
    x = "",
    y = "Odds Ratio"
  ) +
  theme_classic(base_size = 14)

  
```

```{r, codebook}

generate_codebook <- function(df) {
# Extract variable information
variable_names <- names(df)
variable_values <- get_label(df)  # Retrieves variable labels
values <- get_labels(df)          # Retrieves value labels
variable_types <- sapply(df, class)
variable_labels <- sapply(df, function(x) attr(x, "label"))
value_labels <- lapply(df, function(x) attr(x, "labels"))

# Create the codebook dataframe
codebook <- tibble(
  Variable = variable_names,
  Label = variable_values,
  Type = variable_types,
) %>%
  mutate(
    Values = sapply(value_labels, function(labels) {
      if (is.null(labels)) return(NA)  # Handle missing labels
      paste(names(labels), "=", labels, collapse = ", ")
    })
  )

}

codebook_LAPOP <- generate_codebook(LAPOP)

# Export the common_codebook to HTML
codebook_LAPOP %>%
  kbl(caption = "LAPOP Codebook") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  save_kable("lapop_codebook.html")

```

```{r}

library(survey)
library(stringi)


# Ensure municipality codes are formatted correctly
LAPOP_clean <- LAPOP %>%
  mutate(
    wt = as.numeric(wt),  # Ensure weights are numeric
    male = ifelse(q1tb == 1, 1, 0), 
    
    edr = case_when(
      edr == 3 ~ 1,
      edr %in% c(1,2) ~ 0,
      TRUE ~ NA_integer_
    ),  # College educated (to some level)
    
    q10newt = case_when(
      q10newt == 31 ~ 1,  # 0-195
      q10newt == 32 ~ 2,  # 196-350
      q10newt == 33 ~ 3,  # 351-480
      q10newt == 34 ~ 4,  # 481-620
      q10newt == 35 ~ 5,  # 620+
      TRUE ~ NA_real_  # Assign NA for missing values
    ), # Income level (5-point scale)
    
    jc13 = case_when(
      jc13 == 1 ~ 1,
      jc13 == 2 ~ 0,
      TRUE ~ NA_integer_
    ),  # State force justified if corruption

    jc15a = case_when(
      jc15a == 1 ~ 1,
      jc15a == 2 ~ 0,
      TRUE ~ NA_integer_
    ),  # Executive justified in closing the legislature

    vic1ext = case_when(
      vic1ext == 1 ~ 1,
      vic1ext == 2 ~ 0,
      TRUE ~ NA_integer_
    ),  # Victim of crime in the last 12 months

    b2 = case_when(
      b2 %in% c(1,2,3) ~ 1,  # No (Respect for institutions)
      b2 %in% c(5,6,7) ~ 0,  # Yes
      TRUE ~ NA_integer_
    ), 
    
    b4 = case_when(
      b4 %in% c(1,2,3) ~ 1,  # No
      b4 %in% c(5,6,7) ~ 0,  # Yes
      TRUE ~ NA_integer_
    ), # 	Orgullo por el sistema político
    
    b21 = case_when(
      b21 %in% c(1,2,3) ~ 1,  # No 
      b21 %in% c(5,6,7) ~ 0,  # Yes
      TRUE ~ NA_integer_
    ) # 	Confianza en los partidos políticos
  ) %>%
  select(
    prov1t, # Departamento
    municipio1t, # Municipio
    wt, 
    strata,
    
    male,   
    edr,     # College educated (to some level)
    q10newt, # Income level (5-point scale)
    
    jc13, # Force by the state is justified when there is a lot of corruption
    jc15a, # The executive is justified in closing the legislature
    vic1ext, # Victim of crime in the last 12 months
    b2, # Respect for institutions
    b4, # Pride in political system
    b21 # Confidence in political parties
  ) %>%

# Keep relevant columns
  filter(!is.na(municipio1t) & !is.na(wt))  # Remove rows with missing municipality codes or weights

# Ensure weight variable has no missing values
LAPOP_clean <- LAPOP_clean %>%
  filter(!is.na(wt) & wt > 0)

# Define survey design (accounting for stratification & weights)
survey_design <- svydesign(
  id = ~1,  # No clustering at this level (if there's a PSU variable, replace 1 with it)
  strata = ~strata,  # Use correct stratification variable (modify if needed)
  weights = ~wt,  # Apply probability weights
  data = LAPOP_clean
)

# Compute weighted means & counts by municipality
muni_LAPOP_weighted <- LAPOP_clean %>%
  group_by(municipio1t) %>%
  summarise(
    pct_male = sum(male * wt, na.rm = TRUE) / sum(wt, na.rm = TRUE) * 100,
    n_male = sum(!is.na(male)),  

    pct_edr = sum(edr * wt, na.rm = TRUE) / sum(wt, na.rm = TRUE) * 100,
    n_edr = sum(!is.na(edr)),  

    avg_income_lvl = sum(q10newt * wt, na.rm = TRUE) / sum(wt, na.rm = TRUE),  # Income as continuous
    n_income = sum(!is.na(q10newt)),  

    pct_force_justified = sum(jc13 * wt, na.rm = TRUE) / sum(wt, na.rm = TRUE) * 100,
    n_force_justified = sum(!is.na(jc13)),

    pct_exec_close_leg = sum(jc15a * wt, na.rm = TRUE) / sum(wt, na.rm = TRUE) * 100,
    n_exec_close_leg = sum(!is.na(jc15a)),

    pct_vic_crime = sum(vic1ext * wt, na.rm = TRUE) / sum(wt, na.rm = TRUE) * 100,
    n_vic_crime = sum(!is.na(vic1ext)),

    pct_respect_inst = sum(b2 * wt, na.rm = TRUE) / sum(wt, na.rm = TRUE) * 100,
    n_respect_inst = sum(!is.na(b2)),

    pct_pride_political_system = sum(b4 * wt, na.rm = TRUE) / sum(wt, na.rm = TRUE) * 100,
    n_pride_political_system = sum(!is.na(b4)),

    pct_confidence_parties = sum(b21 * wt, na.rm = TRUE) / sum(wt, na.rm = TRUE) * 100,
    n_confidence_parties = sum(!is.na(b21)),

    n_total = sum(!is.na(male) | !is.na(edr) | !is.na(q10newt) | !is.na(jc13) | !is.na(jc15a) | !is.na(vic1ext) | !is.na(b2) | !is.na(b4) | !is.na(b21))
  ) %>%
  ungroup()


```

```{r}
# Compute unweighted means & counts by municipality
muni_LAPOP <- LAPOP_clean %>%
  group_by(municipio1t) %>%
  summarise(
    pct_male = ifelse(sum(!is.na(male)) == 0, NA, mean(male, na.rm = TRUE) * 100),
    n_male = sum(!is.na(male)),  

    pct_edr = ifelse(sum(!is.na(edr)) == 0, NA, mean(edr, na.rm = TRUE) * 100),
    n_edr = sum(!is.na(edr)),  

    avg_income_lvl = ifelse(sum(!is.na(q10newt)) == 0, NA, mean(q10newt, na.rm = TRUE)),  # Income as continuous
    n_income = sum(!is.na(q10newt)),  

    pct_force_justified = ifelse(sum(!is.na(jc13)) == 0, NA, mean(jc13, na.rm = TRUE) * 100),
    n_force_justified = sum(!is.na(jc13)),

    pct_exec_close_leg = ifelse(sum(!is.na(jc15a)) == 0, NA, mean(jc15a, na.rm = TRUE) * 100),
    n_exec_close_leg = sum(!is.na(jc15a)),

    pct_vic_crime = ifelse(sum(!is.na(vic1ext)) == 0, NA, mean(vic1ext, na.rm = TRUE) * 100),
    n_vic_crime = sum(!is.na(vic1ext)),

    pct_respect_inst = ifelse(sum(!is.na(b2)) == 0, NA, mean(b2, na.rm = TRUE) * 100),
    n_respect_inst = sum(!is.na(b2)),

    pct_pride_political_system = ifelse(sum(!is.na(b4)) == 0, NA, mean(b4, na.rm = TRUE) * 100),
    n_pride_political_system = sum(!is.na(b4)),

    pct_confidence_parties = ifelse(sum(!is.na(b21)) == 0, NA, mean(b21, na.rm = TRUE) * 100),
    n_confidence_parties = sum(!is.na(b21)),

    n_total = sum(!is.na(male) | !is.na(edr) | !is.na(q10newt) | !is.na(jc13) | !is.na(jc15a) | 
                  !is.na(vic1ext) | !is.na(b2) | !is.na(b4) | !is.na(b21))
  ) %>%
  ungroup()

```

```{r}

# Extract municipality and department mappings from `codebook_LAPOP`
location_df <- codebook_LAPOP %>%
  filter(Variable %in% c("municipio1t", "prov1t")) %>%  # Select municipio1t & prov1t
  select(Variable, Values) %>%  # Extract relevant columns
  separate_rows(Values, sep = ", ") %>%  # Split each mapping into its own row
  mutate(
    location_name = str_extract(Values, "^[^=]+"),  # Extract name before "="
    location_code = str_extract(Values, "(?<=\\= )\\d+")  # Extract numeric code after "="
  ) %>%
  select(Variable, location_code, location_name) %>%
  mutate(location_code = as.character(location_code))  # Convert codes to character for merging

# Create separate dataframes for municipalities and departments
muni_df <- location_df %>% 
  filter(Variable == "municipio1t") %>% 
  select(-Variable) %>% 
  rename(municipio1t = location_code, municipality_name = location_name)

dept_df <- location_df %>% 
  filter(Variable == "prov1t") %>% 
  select(-Variable) %>% 
  rename(prov1t = location_code, department_name = location_name)

# Convert codes to character type in LAPOP for correct merging
LAPOP <- LAPOP %>%
  mutate(
    municipio1t = as.character(municipio1t),
    prov1t = as.character(prov1t)
  ) %>%
  left_join(muni_df, by = "municipio1t") %>%  # Merge municipality names
  left_join(dept_df, by = "prov1t")  # Merge department names


LAPOP2 = LAPOP %>%
  select(department_name, prov1t, municipality_name, municipio1t) %>%
  distinct(municipio1t,.keep_all = TRUE) %>%
  rename(municipio = municipality_name,
         departamento = department_name)

library(stringi) 

LAPOP2 = LAPOP2 %>% 
  mutate(municipio = str_trim(tolower(stri_trans_general(municipio, "Latin-ASCII"))),
         departamento = str_trim(tolower(stri_trans_general(departamento, "Latin-ASCII"))))

LAPOP2$municipio[LAPOP2$municipio == "nahulingo"] = "nahuilingo"
LAPOP2$municipio[LAPOP2$municipio == "san jose"] = "san jose las fuentes"
LAPOP2$municipio[LAPOP2$municipio == "san antonio"] = "san antonio del mosco"
LAPOP2$municipio[LAPOP2$municipio == "oscicala"] = "osicala"

# Check if values now match
intersect(clean_merged$municipio, LAPOP2$municipio) %>% length()

merged_check = clean_merged %>%
  select(departamento, municipio)

missing_in_LAPOP2 <- merged_check %>%
  anti_join(LAPOP2, by = "municipio") %>%
  distinct(municipio, .keep_all = TRUE)

missing_in_merged <- LAPOP2 %>%
  anti_join(merged_check, by = "municipio") %>%
  distinct(municipio, .keep_all = TRUE)

merged_one <- left_join(clean_merged, LAPOP2, by = c("departamento" = "departamento", 
                                                     "municipio" = "municipio"))

```

```{r}

merged_one = merged_one %>%
  mutate(municipio1t = as.double(municipio1t))

merged = left_join(merged_one, muni_LAPOP, by = "municipio1t")

saveRDS(merged, "finalproject/01_data/clean_output.rds")
```

```{r}
clean_merged = readRDS("finalproject/01_data/clean_output.rds")
names(clean_merged)

# Run the OLS regression
model_bukele <- lm(n_share ~ avg_income_lvl +
                     pct_force_justified + pct_exec_close_leg + pct_vic_crime + 
                     pct_respect_inst + pct_pride_political_system + pct_confidence_parties + 
                     densidad + percent_urbano + masculinidad + relacion_dependencia + 
                     x60_anos_y_mas + tgf + tmi + tasa_analfabetismo + asistencia_escolar +
                     agua_potable + electricidad + sin_servicio_sanitario + con_piso_de_tierra,
                   data = clean_merged)

model_simplified <- lm(n_share ~ 
                         avg_income_lvl + percent_urbano + tasa_analfabetismo + 
                         sin_servicio_sanitario +
                          masculinidad +
                         pct_force_justified + pct_exec_close_leg + pct_vic_crime,
                       data = clean_merged)

model3 <- lm(n_share ~ 
                         avg_income_lvl + percent_urbano + tasa_analfabetismo + 
                        masculinidad +
                         pct_force_justified + pct_exec_close_leg + pct_vic_crime,
                       data = clean_merged)

library(stargazer)
stargazer(model_bukele, model_simplified, model3, type = "text")

library(car)
vif(model_bukele)
vif(model_simplified)


```

```{r}

model1 <- lm(n_share ~ pct_vic_crime, data = clean_merged)

# Model 2: Adding demographic and economic controls (including avg_ideology for future me)
model2 <- lm(n_share ~ pct_vic_crime + percent_urbano + avg_income_lvl + pct_edr, 
             data = clean_merged)

model3 = lm(n_share ~ violent_event, data = clean_merged)

model4 <- lm(n_share ~ violent_event + pct_vic_crime + percent_urbano + avg_income_lvl + pct_edr, 
             data = clean_merged)

stargazer(model1, model2, model3, model4, type = "text",
          title = "Regression Results: Violent Crime and Bukele Support",
          dep.var.labels = "Nueva Ideas Vote Share (%)",
          digits = 3)

# Model 3: Adding economic perceptions (mediation test) while controlling for avg_ideology
model3 <- lm(pct_trump ~ pos_change_gdp + avg_ideology + pct_collgrad + avg_income + pct_white + pct_romney + 
               pct_pos_perception_econ, data = county_election)

# Model 4: Including interaction effects for ideology (moderation test)
model4 <- lm(pct_trump ~ pos_change_gdp * avg_ideology + pct_collgrad + avg_income + pct_white + pct_romney + pct_pos_perception_econ * avg_ideology, data = county_election)

# Output regression results in a table
stargazer(model1, model2, model3, model4, type = "text",
          title = "Regression Results: GDP Growth and Trump Support",
          dep.var.labels = "Trump Vote Share (%)",
          digits = 3)
```

```{r}
library(ggdag)
library(dagitty)
library(ggplot2)


dag_model <- dagify(
  Vote_Share ~ Political_Trust + Political_Satisfaction + State_Force_Justified + 
               Illiteracy + Poverty + Urbanization + Infrastructure,
  Political_Trust ~ Illiteracy + Urbanization,
  Political_Satisfaction ~ Political_Trust + Infrastructure,
  Poverty ~ Illiteracy + Urbanization + Infrastructure,
  Infrastructure ~ Urbanization,
  exposure = "Illiteracy",
  outcome = "Vote_Share",
  labels = c(
    Vote_Share = "Bukele's Vote Share",
    Political_Trust = "Trust in Institutions",
    Political_Satisfaction = "Satisfaction with Political System",
    State_Force_Justified = "Belief State Force is Justified",
    Illiteracy = "Illiteracy Rate",
    Poverty = "Poverty Level",
    Urbanization = "Urbanization",
    Infrastructure = "Access to Infrastructure"
  )
)

ggdag_status(dag_model, use_labels = "label", text = FALSE) +
  guides(fill = "none", color = "none") +
  theme_dag()

```
