---
title: "Democratic Backsliding in El Salvador"
subtitle: "POL 194H: Honors Thesis"
author: "Luca Alioto"

format:
  pdf:
    documentclass: article
    pdf-engine: pdflatex
    fontsize: 12pt
    geometry:
      - margin=1in
    include-in-header:
      text: |
        \usepackage{tikz}
        \usetikzlibrary{calc, arrows.meta, positioning, shapes}
        \usepackage{caption}
        \usepackage{fancyhdr}
        \usepackage{titling}
        \usepackage{mathptmx}
        \usepackage{float}
        
        \setlength{\parindent}{2em}   % Controls indent size (standard is ~1.5–2em)
        % Hanno didnt use so in wokring paper so ig I wont as well
        % \usepackage{indentfirst}     % Also indent first paragraph after headings
        
        % --- REQUIRED FOR MANUALLY FIXING SPACING ---
        \usepackage{setspace} 
        \doublespacing       % Starts the document in double spacing
        % Fix double-indented lists
        \usepackage{enumitem}
        \setlist[itemize]{leftmargin=*}
        \setlist[enumerate]{leftmargin=*}
        % --------------------------------------------

        % ---- Fancy header setup ---- 
        \fancypagestyle{plain}[fancy]{}
        \fancyhf{}
        \fancyhead[L]{Alioto | POL 194HA}
        \fancyhead[C]{}
        \fancyhead[R]{\thepage}
        \renewcommand{\headrulewidth}{0.4pt}

        % ---- Center the Title/Author nicely ----
        \pretitle{\begin{center}\LARGE}
        \posttitle{\par\end{center}\vskip 0.5em}
        \preauthor{\begin{center}\large}
        \postauthor{\par\end{center}\vskip 1em}

execute:
  echo: false
  warning: false
  message: false

bibliography: references.bib
---

```{r setup, include=FALSE}
library(knitr)
library(here)
## opts_knit$set(root.dir = here::here("01_thesis_replication"))
```

```{r, include=FALSE}
# Load necessary libraries
library(tidyverse)
library(dplyr)
library(labelled)
library(kableExtra)
library(sjlabelled)
library(stringr)
library(haven)
library(readxl)
library(stargazer)
library(broom)
library(gt)

LAPOP <- haven::read_dta(here("01_data/02_lapop/SLV_2021_LAPOP.dta")) # Public Opinion data from 2021

LAPOP_A = LAPOP %>%
  filter(core_a_core_b == "Core A") # Group A respondents
```

```{r}

# Helper Function to remove LAPOP "Junk" codes
clean_lapop <- function(x) {
  case_when(
    x %in% c(888888, 988888, 999999) ~ NA_real_,
    TRUE ~ as.numeric(x)
  )
}

LAPOP_Model <- LAPOP_A %>%
  select(
    m1,       # B: DV - Bukele Approval
    vicbar4a, # Ex: IV - Extortion
    idio2,    # Ec: Moderator - Econ Perception
    edr,       # Inc (Proxy): Education Years
    ur1new,   # U: Confounder - Urban
    exc7,     # C: Control - Corruption Perception
    exc6,
    sgl1      # W (Proxy): Municipal Services
  ) %>%
  
 
  mutate(across(everything(), clean_lapop)) %>%
  
  # Step B: Create Analysis Variables
  mutate(
    # --- DV: Bukele Support (B) ---
    # 1=Support (Very Good/Good), 0=No Support (Regular/Bad/Very Bad)
    bukele_support = case_when(
      m1 %in% c(1, 2) ~ 1,
      m1 %in% c(3, 4, 5) ~ 0,
      TRUE ~ NA_real_
    ),
    
    # --- IV: Extortion Victim (Ex) ---
    # Original: 1=Yes, 2=No -> Recode to 1=Yes, 0=No
    extortion_victim = case_when(
      vicbar4a == 1 ~ 1,  # Yes
      vicbar4a == 0 ~ 0,  # No (The Codebook Value)
      TRUE ~ NA_real_
    ),
    
    # --- Moderator: Econ Perception (Ec) ---
    # Original: 1=Better, 2=Same, 3=Worse
    # REVERSE CODE: 3=Better, 2=Same, 1=Worse 
    # (Higher number = Better economy, making the interaction intuitive)
    econ_perception = case_when(
      idio2 == 1 ~ 3,
      idio2 == 2 ~ 2,
      idio2 == 3 ~ 1,
      TRUE ~ NA_real_
    ),
    
    # --- Proxy for Income (Inc): Education ---
    # Continuous variable (Years of schooling)
    
    # --- Confounder: Urban (U) ---
    # 1/2 = Urban, 3/4 = Rural -> Recode to 1=Urban, 0=Rural
    is_urban = case_when(
      ur1new %in% c(1, 2) ~ 1,
      ur1new %in% c(3, 4) ~ 0,
      TRUE ~ NA_real_
    ),
    
    # --- Control: Corruption Perception (C) ---
    # Original: 1=Very Common ... 4=Uncommon
    # REVERSE CODE: 4=High Corruption ... 1=Low Corruption
    corruption_perc = case_when(
      exc7 == 1 ~ 4,
      exc7 == 2 ~ 3,
      exc7 == 3 ~ 2,
      exc7 == 4 ~ 1,
      TRUE ~ NA_real_
    ),
    
    edr = edr, # 0 as none 3 as Uni 
    bribe_victim = exc6, 
    
    # --- Proxy for Welfare (W): Muni Services ---
    muni_services = case_when(
      sgl1 == 1 ~ 5,
      sgl1 == 2 ~ 4,
      sgl1 == 3 ~ 3,
      sgl1 == 4 ~ 2,
      sgl1 == 5 ~ 1,
      TRUE ~ NA_real_
    ),
  )

```

## Introduction

Some believe that the canonizing event in the democratic backsliding of El Salvador began on March 26th, 2022, a day that saw the highest mortality rate in two decades. In the span of 24 hours, sixty-two people were killed in a gang attacks; a violent breakdown in government-gang negotiations [@gellman2022]. 81st President Nayib Bukele responded by declaring a “war on gangs,” establishing a State of Exception that suspended the constitutional rights to freedom of assembly, privacy in communications, and due process. In the weeks that followed, military officials were deployed in the streets, and authorities arrested over 8,500 people—a number that would increase to nearly 80,000, or more than 1% of the population, by 2024. As Bukele's response caused widespread international condemnation and concerns over arbitrary detention, his domestic approval rating soared to nearly 90%, showing that his iron fist approach to gangs and gang violence was backed by the citizens of El Salvador [@gellman2022; @meléndez-sánchez2024].

While this marks an escalation in Bukele's centralization of power this was merely the dramatic display of the erosion. Research from scholars and international observers suggests that it began much earlier, when his administration worked to systematically undermine opposition parties and lower the efficacy of courts, independent agencies, local governments, civil society organization, and the press. Scholars argue the country’s authoritarian shift can be marked by the rapid acceleration in the centralization of power, occurring after his party, Nuevas Ideas, secured a legislative supermajority in 2021 [@meléndez-sánchez2022]. Following the 2021 legislative elections, the Assembly swiftly removed all five magistrates of the Constitutional Chamber of the Supreme Court and the Attorney General, replacing them with loyalists [@u.s.departmentofstateHumanRightsReports2021; @gellman2022]. Later that year, the reconstituted court authorized presidential reelection despite explicit constitutional prohibitions, paving the way for Bukele’s successful 2024 bid which resulted in his re-introduction to a government that had been radically reshaped [@bbc2024].

Long before Bukele entered office, Salvadorans had grown deeply disillusioned with representative democracy. In 2018, the year before his election, 63.4 percent reported dissatisfaction with democracy [@lapoplabAmericasBarometerSalvador2018]. More than 60 percent agreed that elections were “a waste of time,” and nearly 80 percent said major parties did not represent people like them [@iudop2018; @meléndez-sánchez2022]. Support for traditional parties had collapsed: just 30.8 percent identified with two of the promenient political parites by 2018. Public distrust extended further, as 84.9 percent of Salvadorans believed that at least half of all politicians were corrupt [@lapoplabAmericasBarometerSalvador2018]. In addition to perceptions of widespread corruption, Salvadorans faced large levels of insecurity and exposure to violence. Between 2004 and 2018, crime and violence consistently emerged as the country’s most severe problem, rising from 32 to 62 percent in national opinion surveys. Even during periods of declining homicide rates—such as the 2012–2014 truce—between 15 and 20 percent of Salvadorans reported living in neighborhoods controlled by gangs [@lapoplabAmericasBarometerSalvador2018; @castro].

Previous governments failed to contain violence and corruption, producing a widening gap between Salvadorans and the country’s democratic institutions. Bukele filled that void through appealing to widespread frustration and insecurity [@vanderborgh2019; @humanrightswatch2023]. He won 53 percent of the vote—21 points ahead of his nearest competition [@márquez2021]. His enduring popularity had gone beyond the campaign trail, steming not merely from rhetoric but from his ability to deliver on it. Although the March 2022 security crackdown generated significant human-rights concerns, it achieved what previous administrations were unable to. By early 2023 the gangs that had long dominated territory, extorted residents, and driven El Salvador’s status as one of the world’s most violent countries had been largely dismantled. Official homicide and extortion rates fell to historic lows. While Bukele is not the first Latin American president to adopt iron fist policies, he is the first to do so with outcomes that, from the public’s perspective, appeared to decisively resolve the very insecurities that had eroded trust in the democratic system [@meléndez-sánchez2022].

Bukele has roooted his political project in the claim that he, not existing institutions, embodies the true will of the Salvadoran people. He capitalized on widespread anger toward corruption with slogans such as “Return what you have stolen” and “There is enough money when no one steals". In speeches, he routinely frames his administration as a historic effort to restore “real democracy” after two centuries of elite manipulation and institutional decay. “Now we are building a real democracy,” he proclaimed. “For 200 years, democracy was a pantomime… They never cared about people, they only cared about votes” [@meléndez-sánchez2022]. Following his removal of term limits and subsequent re-election, Bukele went further, asserting that democratic legitimacy derived exclusively from his overwhelming popular mandate noting,

> “This is the first time that a single party exists in a country in a fully democratic system. The entire opposition was pulverized… Democracy means the power of the people, demos and kratos. We Salvadorans are united,”

\noindent labeling his consolidation of power and removal of both of the longstanding major parties from participation as democratic [@meléndez-sánchez2022; @wolf2024]

Largely, public opinion trends suggest that many Salvadorans have come to accept this reframing. Between 2018 and 2021, satisfaction with democracy rose sharply, and distrust in elections declined. In effect, Salvadorans have reported growing confidence in what they perceive to be a democratic system at the very moment that core democratic norms and institutions are being weakened[@lapoplabAmericasBarometerSalvador2018; @meléndez-sánchez2023].

As perceptions of democracy concentrate around the executive, and his approach to creating "the safest country in North America," understanding the mechanisms sustaining such high support is critical [@velez2024]. This raises a fundamental question for the study of democratic backsliding in contexts of high violence:

> **How does exposure to state and criminal violence influence support for President Bukele and authoritarian policies in El Salvador?**

\noindent This study examines whether a dynamic of performance legitimacy and selective accountability sustains Bukele's popularity during this period of democratic erosion. By analyzing individual-level data from the 2021 AmericasBarometer (LAPOP), I explore the association between exposure to violence—both criminal and state-led—and support for the incumbent. As well, I work to understand the effect economic perceptions and the expansion of Bukele's welfare state act as compensatory mechanisms, potentially overriding the political costs of security failures.

### Research Design

To answer this question, I utilize a comparative Logistic Regression design on a national probability sample of Salvadoran adults ($N=1,435$). I operationalize victimization in two distinct ways—Gang Extortion (representing policy failure) and State Bribery (representing institutional abuse)—to test if voters attribute blame differently based on the perpetrator. To isolate these mechanisms, I control for structural confounders including urbanization and education (serving as a proxy for socioeconomic status) and test for the moderating role of retrospective economic perception.

### Key Findings

In this paper, I find that victims crime do not blindly demand authoritarianism or call for iron fist policies. Instead, my results reveal a process of selective accountability. I find that victims of gang extortion are significantly less likely to support President Bukele ($\beta = -0.465$), treating continued insecurity as a specific policy failure. However, voters do not penalize the administration for state-led corruption, suggesting Bukele is largley insulated from the reputational costs of institutional abuse. Ultimately, I find that economic performance serves as a critical buffer, with positive economic perceptions offsetting the disapproval generated by security failures, effectively buying the political capital necessary to weather the costs of violence.

## Literature Review

### Civil War, the Peace Accords, and the Long Term Institutional Effects

While many foreign observers initially regarded the Salvadoran Peace Accords of 1992 as Latin America’s most successful democratic transitions, scholars have since identified the pact as a source of institutional instability and public disillusionment with democracy[@buiza2018; @meléndez-sánchez2021]. These negotiations preceded a brutal civil war between the insurgent Farabundo Martí National Liberation Front (FMLN) and the Salvadoran government, administered in its final years by the right-wing Nationalist Republican Alliance (ARENA). Throughout the twelve-year period, the conflict claimed the lives of 75,000 civilian and thousands of combatants alongside displacing forcefully displaced nearly one million people [@chávez2015].

The civil war itself was rooted in decades of extreme socioeconomic inequality, elite domination, and systematic state repression under what critics label authoritarian military-oligarchic regime [@chávez2015]. In this context, the FMLN—a coalition of leftist guerrilla organizations—emerged with the objective of overthrowing the authoritarian state and fundamentally redistribute political and economic power. The founding of the FMLN in 1980 represented a pivotal convergence of the “Old Left” and the “New Left,” creating a group with substantial organizational capacity, military and political expertise, and international networks. While this ideological and strategic diversity was the groups initial strength, throughout the conflict leaders struggled to maintain unified positions on revolutionary strategy, negotiation, and long-term political goals. At the same time, the ideological decline of socialist parties in Latin America and Europe pushed the FMLN toward a more pragmatic orientation and away from its origional rigid Marxist-Leninist principles [@chávez2015; @meléndez-sánchez2021].

This consestions largely influenced the peace negotiations in 1992, where the organization set aside its earlier insistence on structural socioeconomic transformation. Instead, FMLN leadership articulated two core negotiating aims: the comprehensive demilitarization of the Salvadoran state and the creation of institutional guarantees for a competitive democratic order. The UN-mediated negotiation between the government of President Alfredo Cristiani and the FMLN ended the conflict in 1992, paving the way for the only sustained democratic period in Salvadoran [@chávez2015].

In their paper, @meléndez-sánchez2022 highlights three central components of the transition pact that initially enabled the emergence of a strong and stable party system following the peace accords. The first was the peace accords themselves, which transformed the FMLN from a mobilized guerrilla force into a political party able to participate freely in the political sphere. Second, the 1993 General Amnesty Law granted combatants on both sides immunity from prosecution for war-related crimes, allowing the leadership to remain in power after the war. Finally, a new Electoral Code, written in 1992, established high barriers to entry for new parties and gave leaders of both ARENA and the FMLN significant influence over future electoral processes.

Scholarship argues that these “pacted transitions,” aimed at safeguarding the interests of powerful elites, are key to stabilizing new democracies and creating “a sufficiently strong consensus about the rules of the game” so that no major elite actor is incentivized to revert to authoritarianism [@karl1990; @meléndez-sánchez2022]. This approach to building state capacity rests on the idea that conflict is rooted in weak institutional frameworks, leading external actors to focus their efforts on strengthening institutional capacity through state-building and democracy-support measures [@fox2012; @zulueta-fülscher2018]. The three components of El Salvador’s transition followed this logic, successfully persuaded the parties to lay down their arms, created conditions “minimally safe for powerful actors on both sides of the regime divide,” and enabled two historical rivals to form a democratic system [@meléndez-sánchez2022].

Nonetheless, scholarship suggests that when capacity-building becomes the primary focus of development plans, elites often sideline efforts to make institutional frameworks more inclusive or democratic. Increaing representation, participation, openness, and accountability would require structural reforms that depend on elites relinquishing at least part of their power to either the public or emerging political actors [@zulueta-fülscher2018]. More broadly,

> *“*there is no necessity for a society to develop or adopt the institutions that are best for economic growth or the welfare of its citizens, because other institutions may be even better for those who control politics and political institutions*”* [@daronacemoglu2013].

\noindent with being elites likely to resist and reforms which reduce their institutional control [@zulueta-fülscher2018].

Over time, the high barriers to entry and protections for existing leadership built into the 1992 peace accords encouraged a form of democratic consolidation in which a small group of elites on both sides of the regime divide could reliably safeguard their interests through ARENA and the FMLN. A key factor was the Amnesty Law, which allowed wartime elites to remain in control of both parties and prevented perpetrators of war crimes and corruption from being held accountable. Additionally, the Electoral Code gave party leaders full control over internal nomination and discipline, enabling these elites to remain in power for decades while removing any emerging opposition [@meléndez-sánchez2022]. This control extended across the broader political system. @gutiérrezsalazar2015 notes that postwar reforms did little to limit partisan influence over the judiciary, pointing to limited autonomy and extensive clientelist networks as key impediments to the maintenance of democratic norms.

These arrangements also preserved and deepened longstanding socioeconomic and political inequalities, allowing various forms of elite domination to persist and evolve in the following decades. As a result, inequality, economic exclusion, and social disparities remained entrenched, driving increases in external economic institutions through migration and and a large dependence on remittances.

### Visible Corruption, Democratic Consolidation, and Narrow Consensus

Democratic consolidation also produced a narrow procedural consensus, marked by limited ideological diversity between parties. In their papers, @colalongo2024 and @meléndez-sánchez2023 point to the 2009 election as a key example of this dynamic. After twenty years of ARENA governments, the FMLN won the presidency for the first time since the democratic transition and formed a left-wing administration under Mauricio Funes, a political outsider to the party. Funes’s election was widely interpreted as a compromise, signaling that the FMLN had sacrificed ideological purity to win the presidency for the first time. In hopes of curtailing this and retaining internal power, party leadership had initially vetoed Funes' nomination, later permitting him to run only on the condition that his running mate be a former FMLN civil war commander.

Despite his reformist platform, Funes largely replicated the political model of his ARENA predecessors, and party brands began to dilute. The FMLN, whose identity had long been built around opposing ARENA’s economic and security policies, ultimately favored policy continuity over reform. Contrary to campaign promises, the party did not attempt to de-dollarize the economy, nationalize key industries, or intensify redistribution; instead, Funes instituted a government austerity plan in his first month in office [@meléndez-sánchez2023]. Economically, ongoing market liberalization continued to erode state institutions, sustaining structural poverty and contributing to social violence [@colalongo2024].

This dilution extended to security policy. After experimenting with alternative approaches, Funes ultimately embraced the hardline gang crackdowns of previous ARENA governments. Funes’s decision to militarize public security in response to renewed gang offensives further entrenched this overlap, paving the way for deeper institutional violence [@roque2021; @colalongo2024]. This was only intensified after Funes was succeeded by Sánchez Cerén, who, like his predecessor, continued to favor free-market policies and repressive security approaches [@meléndez-sánchez2023].

In this environment, civic–military alliances persisted to preserve the status quo: elites protected their interests, constrained reforms, and prevented substantive debate over the country’s economic model. @chávez2015 notes that the FMLN’s turn toward pragmatism—and the concessions made to consolidate itself as a legal political party—anchored it firmly within ARENA’s neoliberal framework rather than its original transformative project. The presidency of Mauricio Funes thus marked an institutional shift away from the FMLN’s socio-economic reform agenda, solidifying its role as a center-left party that had conceded the economic status quo.

Alongside brand dilution, corruption investigations involving members of both ARENA and the FMLN reinforced voters’ perceptions that the two parties had become increasingly indistinguishable in their policy orientations and political behavior. A major catalyst for this disillusionment was a string of high-profile corruption cases beginning in 2013.

In September of that year, former President Francisco Flores (1999–2004) was accused of redirecting \$15 million in international earthquake-relief donations to ARENA’s campaign coffers. What followed was unprecedented: within a five-year span, six of the country’s most powerful political figures were publicly and credibly implicated in corruption. While in theory, high-profile corruption prosecutions can strengthen democracy by demonstrating institutional accountability and deterring future abuses, they can also backfire by generating a broader perception that the entire political system is corrupt. Scholars describe this as a sense that “the whole system is rotten,” a dynamic which proved especially destabilizing, as it coincided with the collapse of meaningful programmatic differences between ARENA and the FMLN. The result was a voter base increasingly disillusioned with the existing democratic structure which seemed aligned in their ideological and material goals [@meléndez-sánchez2022] .

These instances of corruption and alignment were not merely a matter of policy missteps but reflected a deeper structural feature of the postwar regime, one fundamentally centered on maintaining order rather than expanding representation or accountability. El Salvador’s democracy became, in effect, “hollowed out” by design— providing populist candidates little push back when attacking a democracy that failed to move beyond capacity-building.

### Gangs, Mano Duro, and the Future for Bukele

The roots of El Salvador’s two most dominant gangs lie in Los Angeles, California, where they were created by the children of asylum seekers and refugees displaced by the country’s civil war. Both Mara Salvatrucha (MS-13) and the 18th Street gang (M-18) emerged as youth responses to exclusion and discrimination, offering protection from rival Mexican-American and African American gangs. Following the end Salvadoran and Guatemalan civil wars, and the passage of the U.S. Illegal Immigration Reform and Immigrant Responsibility Act led to the mass deportation of residents with criminal records. These deportations effectively exported American gang structures to Central America, accelerating the spread of MS-13 and M-18 throughout the region [@fogelbach; @olate2012].

Research on gang membership highlights the social and demographic conditions that shape participation. @olate2012 finds that most gang members are male, unemployed, live in urban areas, and have dropped out of school. Their activities tend to be localized and small-scale, involving drug use and minor trafficking, petty theft, muggings, extortion of local businesses and public transportation, territorial disputes, weapons violations, and violence against rival gang members. Although their presence was significant, gangs did not fully account for the country’s broader violence: evidence from the late 2000s suggested no involvement in large-scale transnational drug trafficking [@undoc2010], and MS-13 and M-18 were responsible for less than a third of homicides and extortions [@olate2012].

Scholars widely agree that the large-scale escalation of gang presence and violence in the 2010s was driven in part by a series of failed punitive approaches. hese measures focused on mass arrests, joint police–military patrols, and broad criminalization of gang membership leading to a dramatic rise in the incarcerated population, especially among gang members [@olate2012]. Across the board the judiciary and prison systems lacked the resources and institutional capacity to process the surge in arrest, with overcrowded prisons becoming incubators for gang organization and recruitment. Additionally, the effect on crime was short-lived: temporary declines were followed by renewed and intensified violence. Repression pushed gangs to become more organized, more territorially coherent, and more violent [@olate2012]. As @wolf2017 notes, governments across the political spectrum pursued *mano dura* largely for political and electoral gain, even as evidence accumulated that it was exacerbating the problem.

Alongside punitive strategies, governments also engaged in informal or covert pacts with gangs—policies that often masked rather than resolved violence. The 2012 truce, one of the most notable examples, rested on secret negotiations that exchanged reduced homicide rates for improved prison conditions and other benefits. When the truce collapsed after roughly a year, violence surged dramatically: the homicide rate reached 106 per 100,000 inhabitants in 2015, the highest in the world at the time [@wolf2024].

Unlike previous administrations, many scholars view Bukele's iron fist policies, most notably the 2022 gang crackdowns, as largely successful. Between 2022 and 2023, the registered homicide rate fell from 8 to 2 per 100,000 inhabitants, the lowest level since the Peace Accords. Field research in ten former gang strongholds suggests that emergency rule has disrupted leadership structures and demobilized much of the rank-and-file. In many historically gang-dominated communities, residents report a sense of safety with large-scale support for iron fist approaches [@wolf2024].

Nonetheless, many scholars are skeptical. The lack of publicly available statistics makes it nearly impossible to verify who is being detained, on what grounds, and their level of gang affiliation. Police reports obtained by InSight Crime show that most detainees are not classified as gang members but as aspiring members or “collaborators,” who tend to be coerced participants . By September 2023, only 42 percent of more than 77,000 detainees were designated as gang members, while 54 percent were labeled collaborators and 4 percent wannabes [@insightcrime; @wolf2024]. Analysts note that dozens of gang cliques remain active and that extortion persists; and while the operate at far smaller scales they still present themselves as a viable option for youth. @wolf2024 argues the emergency regime has targeted young and poor residents of gang-affected neighborhoods raising concerns that the crackdown may entrench the same conditions of poverty, stigma, and coercion that facilitate gang recruitment.

As well, despite Bukel's public condemntation of his predecessor’s negotiations with gangs, many of his approaches mirror those of past iron fist policies. Investigations by the independent outlet *El Faro* revealed that his administration forged an undisclosed pact with gang leaders to secure electoral support and a reduction in violence [@martínez2021]. Notably, negotiations have already sparked conflict, with the mass killings in late March 2022 were being a response to a breakdown between Bukele's administration and gang leadership.

Additionally, Bukele faces the deeper challenge of maintaining the political legitimacy of the emergency regime itself. To sustain popular consent, an autocratic government must simultaneously claim progress and perpetuate the sense of threat that justifies extraordinary measures [@matovski2021; @wolf2024]. Leaders which largely depend on addressing a single-issues to generate public support risk becoming obsolete both when they fail and when they succeed in resolving the crisis that legitimizes their rule. Bukele has attributed recent security improvements to the state of emergency, while also invoking the specter of lingering gang violence to justify its indefinite continuation. This strategy has proven successful, but it requires continually reinforcing the perception of threat and raises questions about how public opinion will respond to any breakdown in the current sense of safety [@wolf2024].

Bukele’s widespread popularity is tightly bound to the state of emergency and the perception that he has decisively confronted gang violence. Outside the realm of security policy, however, the country’s broader conditions mirror many of the failures of past administrations, and in some areas have deteriorated, raising questions around a future where gang policy becomes obsolete.

During his first campaign, Bukele promised to build fifteen major infrastructure projects, including transportation facilities, public schools, and hospitals, at an estimated cost of \$1.5 billion [@labrador2023], many of which have been shelved or postponed [@wolf2024].

In contrast, scholars have criticized Bukele's regime for corruption as well as prioritizing his own political and economic gain. The CECOT facility, as an example, has become central to his internal and external popularity, suggesting a focus on funding that which is politically convenient. Additionally, the administration has used its few finished and many unfinished public works projects toward legitimizing practices of opaque government spending and contracts which ciriiques claim benefit relatives and members of Nuevas Ideas. As well, ministers overseeing the environment, public works, and housing appear to have been selected less for technical expertise than for their loyalty to Bukele and their willingness to circumvent regulatory frameworks [@wolf2024].

Despite the limited public reaction so far, these development projects and land-use practices place pressure on long-standing points of political contention. Throughout the country’s history, land speculation and corruption have been major drivers of public discontent, making Bukele’s current governance strategy a risky gamble.

What makes this approach particularly concerning for Bukele is the country’s socioeconomic conditions remain deeply strained. Almost half the population experiences food insecurity, and between 2019 and 2022, extreme poverty rose by two percentage points to 8.7 percent [@seelke2024]. The economy remains heavily dependent on remittances, which account for 24 percent of GDP, and recorded growth of only 2.6 percent in 2023, making it the least dynamic in Central America [@gonzálezdíaz2024].

Public opinion reflects these pressures. Surveys indicate that Salvadorans are increasingly concerned about unemployment and the rising cost of living rather than security. In 2023, 70 percent of respondents identified economic issues as the nation’s primary problem [@iudop2024]. Currently, the economic stability and growth exists under a model which is unsustainable. While declining crime under the state of emergency has encouraged greater domestic and diaspora investment, foreign direct investment remains low [@seelke2024]. Meanwhile, government debt has surpassed 90 percent of GDP, [@gonzálezdíaz2024] putting El Salvador’s ability to meet its debt obligations as well as its political promises in the coming years into question.

### Gaps in the Literature: Public Opinion and Democratic Perceptions

Despite a growing body of research, important gaps remain in understanding how public opinion responds to state violence and how citizen tolerance for authoritarianism is shaped by their experiences and identities. Much of the existing literature focuses on policy effects or elite behavior, with comparatively little attention to how support for authoritarian regimes is constructed and sustained from below. For example, @castro2022 examine how gang presence affects electoral participation, showing that criminal governance can significantly shape electoral outcomes, yet they do not extend this analysis to perceptions of democratic legitimacy.

Additionally, in their report @meléndez-sánchez2023 shows acts of democratic erosion in El Salvador have had little negative impact on public opinion. In fact, public satisfaction with democracy has increased since Bukele’s election, rising sharply between 2018 and 2021. Paradoxically, Salvadorans have grown more confident in a political system that is, by objective institutional standards, becoming less democratic. Nonetheless, the paper only examines political attitudes toward democracy itself, rather than public opinion and the electoral success of Bukele.

Despite extensive discussion of democratic backsliding, relatively little research examines Bukele himself as the central object of public opinion. There is also limited empirical work on how state violence, notably mass arrests under the state of emergency, shapes voting behavior or long-term attitudes toward authoritarianism. Existing studies do not yet capture whether public support hinges solely on continued security “success,” or whether failures in maintaining safety might prompt voters to look beyond rhetoric. It remains unclear whether Bukele is insulated from or vulnerable to the same patterns of disillusionment that undermined ARENA and the FMLN.

## Causal Pathways & Theoretical Motivation

To examine the association between exposure to violence and economic conditions and support for democratic governance under Bukele, I employ a quantitative research design utilizing survey data from the 2021 LAPOP AmericasBarometer for El Salvador. The analysis focuses on individual-level responses to questions regarding experiences of state violence, perceptions of economic well-being, and attitudes toward democracy.

### Theory: Data-Generating Process (DGP)

To understand the determinants of incumbent support in the context of democratic backsliding, I propose a Data-Generating Process (DGP) where individual political preference ($Y$) is a function of exposure to violence ($X$), conditional on the voter's economic satisfaction ($Z$).

\noindent The causal mechanism linking violence to support for Bukele is conceptualized as follows:

\noindent Exposure to Violence ($X$): Exposure to insecurity serves as the primary independent variable. I theorize that the source of violence determines the political reaction:

-   Gang Extortion ($X_{Gang}$): Following the "Iron Fist" literature, exposure to non-state violence is hypothesized to increase perceptions of insecurity, potentially driving a demand for authoritarian order and increasing support for the strongman ($H_1$).
-   State Predation ($X_{State}$): Conversely, exposure to state-led predation (Bribery) represents a violation of the social contract. This experience is hypothesized to decrease satisfaction with democratic governance and reduce incumbent support ($H_2$).

\noindent Economic Perceptions ($Z$): Economic satisfaction acts as a moderating variable. Individuals with positive economic perceptions are influenced by perceptions of the administration's management of the economy and distribution of economic success. I hypothesize that this "performance legitimacy" weakens the relationship between violence and political preference. Specifically, positive economic evaluations may neutralize the dissatisfaction caused by state violence or amplify the support derived from security demands ($H_3$).

\noindent Support for Bukele ($Y$): The outcome variable represents the individual's binary approval of Bukele’s governance. It is modeled as the net result of these competing pressures: the security experience ($X$) and the economic experience ($Z$).

\noindent Control Variables($\mathbf{C}$):To block confounding paths and isolate the causal effect of violence, I condition on the following structural factors:

-   Urbanization: Urban residents may have distinct exposure levels to extortion networks and possess different baseline political attitudes compared to rural residents.
-   Income (SES): Socioeconomic status influences both the probability of victimization (target suitability) and political preferences. (Operationalized via Education).
-   Perceptions of Corruption: An individual's general view on political corruption is included to distinguish specific victimization events from generalized anti-system cynicism.

\vspace{1em}

These hypothesized relationships are visualized in the Directed Acyclic Graph (DAG) presented in Figure 1. The diagram illustrates the causal pathways linking Victimization ($V$) to Incumbent Support ($B$), while explicitly mapping the structural confounders—such as Urbanization ($U$) and Income ($Inc$)—that simultaneously influence the likelihood of victimization and political attitudes. The dashed line connecting Economic Perception ($Ec$) to the main pathway represents the theorized moderating effect ($H_3$).

\begin{center}
\begin{tikzpicture}[
    % Styles
    var/.style={circle, draw=none, fill=none, inner sep=2pt, font=\large\bfseries},
    direct/.style={->, >=Latex, thick},
    interact/.style={->, >=Latex, thick, dashed}
]

    % --- MAIN AXIS (Centered on 0,0) ---
    \node[var] (v0) at (-3, 0) {V};   % IV: Extortion/Violence
    \node[var] (v1) at (3, 0) {B};     % DV: Bukele Support (Right)

    % --- THE MODERATOR (Centered Top) ---
    \node[var] (v7) at (0, 1) {Ec};    % Ec: Econ Perceptions

    % --- LEFT SIDE (Background/Confounders) ---
    \node[var] (v2) at (-3, 1.5) {U};    % Urban (Above Ex)
    \node[var] (v4) at (-1.5, 2.5) {Inc}; % Income (Higher up, between U and Ec)
    \node[var] (v3) at (-1.5, -1.5) {W};  % Welfare (Below path)

    % --- RIGHT SIDE (Outcomes/Controls) ---
    % REPLACED Id WITH C HERE (Top Right)
    \node[var] (v5) at (3, 1.5) {C};     % Corruption Perception

    % --- ARROWS ---
    
    % 1. The Main Path (Ex -> B)
    \draw[direct] (v0) -- (v1) coordinate[midway] (midpoint);
    
    % 2. The Interaction (Ec -> Center)
    \draw[interact] (v7) -- (midpoint);
    
    % 3. Moderator Direct Effect
    \draw[direct] (v7) -- (v1);

    % --- CONTROL PATHS ---
    
    % Urban (The Confounder)
    \draw[direct] (v2) -- (v0);   % U -> Ex
    \draw[direct] (v2) -- (v7);   % U -> Ec
    \draw[direct] (v2) -- (v5);   % U -> C (Urban affects Corruption Perc)
    \draw[direct] (v2) -- (v4);   % U -> Inc
    
    % Income
    \draw[direct] (v4) -- (v7);   % Inc -> Ec
    \draw[direct] (v4) -- (v5);   % Inc -> C (Income affects Corruption Perc)
    
    % Corruption (Formerly Ideology) -> Bukele
    \draw[direct] (v5) -- (v1);
    
    % Welfare Path
    \draw[direct] (v0) -- (v3);   % Ex -> W
    \draw[direct] (v3) -- (v1);   % W -> B
    
\end{tikzpicture}
\end{center}
\begin{spacing}{1.25}
\begin{tikzpicture}
\node[align=left, font=\small, anchor=west] at (0,0) {
\textbf{Legend} \\
\textbf{V} = Exposure to Victimization (IV) \\
\textbf{B} = Support for Bukele (DV) \\
\textbf{Ec} = Economic Perceptions (Moderator) \\
U = Urbanization \\
Inc = Income \\
C = Perceptions of Corruption \\
W = Exposure to Welfare
};
\end{tikzpicture}
\end{spacing}

### Model Specification

\noindent To test these relationships, I estimate the probability of incumbent support using a Logistic Regression model. The log-odds of support are modeled as:

$$\ln \left( \frac{P(Y_i=1)}{1-P(Y_i=1)} \right) = \beta_0 + \beta_1 X_i + \beta_2 Z_i + \beta_3 (X_i \cdot Z_i) + \mathbf{\gamma C_i} + \epsilon_i$$

\noindent Where:

\begin{spacing}{1.25}
$Y_i$: Binary indicator of Support for Bukele.

$X_i$: Exposure to Violence (modeled separately as $X_{Gang}$ or $X_{State}$).

$Z_i$: Retrospective Economic Perception (Proxy for State Investment).

$X_i \cdot Z_i$: Interaction term testing the compensatory/dampening hypothesis.

$\mathbf{C_i}$: Vector of controls (Urbanization, Education, Corruption Perception, Municipal Services).
\end{spacing}

\noindent Building on this theoretical framework, I test the following hypotheses:

> $H_0$: There is no statistically significant relationship between exposure to victimization and the probability of supporting President Bukele.

#### The "Iron Fist" Hypothesis (Non-State Violence)

\noindent Literature on crime and authoritarianism suggests that high levels of criminal insecurity drive citizens to demand "Mano Dura" (Iron Fist) policies. Under conditions of high gang prevalence, victims are often willing to trade civil liberties and democratic checks for the promise of order. As a result, victimization by non-state actors is expected to increase the appeal of the strongman.

> $H_1$: Individuals exposed to non-state violence (Gang Extortion) will be more likely to support President Bukele, reflecting a demand for authoritarian security.

#### The "Broken Contract" Hypothesis (State Violence)

\noindent On the other hand, direct predication by state actors represents a violation of the social contract no matter the perceived benefits of violence. While citizens may tolerate strict measures against criminals, standard accountability theory suggests they will penalize the incumbent when the state apparatus itself becomes predatory. Therefore, victims of institutional corruption are expected to distinguish between the leader's rhetoric and the state's actual behavior.

> $H_2$: Individuals exposed to state-led predation (Bribery) will be less likely to support President Bukele, reflecting dissatisfaction with institutional corruption.

#### The "Performance Legitimacy" Hypothesis (Moderation)

\noindent Finally, I hypothesize that the political costs of security failures, most notably the disaproval caused by state violence, can be offset by economic performance. Following the logic of "performance legitimacy," perceived economic improvements serve as a compensatory mechanism. I hypothesize that economic satisfaction will moderate the relationship between victimization and support, effectively "dampening" the grievance caused by exposure to violence.

> $H_3$: Positive economic perceptions will weaken the relationship between exposure to victimization and support for President Bukele.

## Data and Measurement

### Unit of Analysis

The unit of analysis for this study is the individual survey respondent, representing the voting-age population (18 years and older) of El Salvador.

### Data Source and Coverage

To examine the small-scale foundations of authoritarian support, I rely on data from the 2021 AmericasBarometer conducted by the Latin American Public Opinion Project (LAPOP). This survey utilizes a stratified, multi-stage cluster probability design to ensure national representatives across both urban and rural strata. The fieldwork was conducted via face-to-face interviews, capturing public sentiment during the consolidation of the Bukele administration but prior to the implementation of the 2022 State of Exception.

### Sample Restrictions

The 2021 AmericasBarometer employs a "split-sample" design (Core A and Core B) to accommodate a broader range of questions. This study utilizes the Core A module, which contains the primary variables regarding victimization and political attitudes. While the total survey sample includes roughly 3,000 respondents, the Core A subset consists of approximately 1,500 individuals. After performing list-wise deletion for missing values on key theoretical variables—specifically victim status and economic perception—the final analytical sample consists of $N = 1,435$ observations.

As well, the analysis excludes responses coded as "Don't Know" $(888888)$, "No Answer" $(988888)$, and "Inapplicable" $(999999)$. These codes represent non-substantive responses that do not map onto the theoretical continua of interest (e.g., the ordinal scale of economic perception or the binary status of victimization). For the "Inapplicable" category, missingness is structural, resulting from survey skip patterns (e.g., questions asked only to specific subsets of respondents). To ensure model validity, these observations are treated as missing data (NA) and removed from the final analytical sample.

### Limitations

While the individual-level approach allows for a direct assessment of the mechanisms linking victimization to political support, the structure of the data limits the analysis to a single point in time. This is especially pertinent given the rapidly evolving political landscape in El Salvador under Bukele, and due to the potential for reverse causality between political attitudes and reported victimization.

As well, the data fails to capture the attitudes of respondents before the 2019 election, where the question of Bukele support would be most relevant. Data from the 2023 LAPOP survey includes a question on vote choice in 2019, but the responses were heavily skewed toward Bukele, with over 70% of respondents reporting voting for him despite only winning around 52% of the vote. This suggests significant social desirability bias in the retrospective vote choice question, and puts into question the validity of using both this retroactive reporting as well as perception variables as a whole as the DV.

Additionally, the split-sample design prevents the entire survey from being utilized, requiring the use of proxies back primarily by theories around the data generating process for certain demographic controls (e.g., Education for Income) that were located in the excluded survey module.

### Key Variables and Coding

This study focuses on the following key variables from the LAPOP 2021 Core A survey and are operationalized as followed:

\noindent Dependent Variable ($Y$): Support for Bukele - Measured as a binary indicator of job approval for President Bukele. Derived from variable $M1$ ("Job Approval") in the 2021 LAPOP survey. To provide a rigorous test of committed support, neutral responses are grouped with the opposition:

\begin{spacing}{1.25}
\begin{equation}
\text{Support for Bukele} =
\begin{cases}
1, & \text{if respondent rates performance as "Good" or "Very Good"} \\
0, & \text{if respondent rates performance as "Fair", "Bad", or "Very Bad"}
\end{cases}
\end{equation}
\end{spacing}

\vspace{1em}

The decision to group neutral responses ("Fair") with the opposition reflects two considerations specific to the Salvadoran context. Methodologically, given the "ceiling effect" of President Bukele's super-majoritarian approval ratings in 2021, treating neutral responses as support would artificially inflate the dependent variable, reducing variance and statistical leverage.

Theoretically, literature on survey behavior in backsliding democracies suggests that "neutral" categories often serve as a "safe harbor" for critical respondents subject to social desirability bias or fear of reprisal [@brownback; @tannenberg2017]. Therefore, coding "Fair" as non-support provides a conservative test of committed incumbent support, distinguishing explicit support from ambivalent and biased responses.

The independent variable of interest is exposure to victimization, operationalized in two distinct forms to capture the dual sources of violence in El Salvador:

\noindent Independent Variable 1 ($X_{Gang}$): Measured as exposure to non-state extortion (Gang Extortion) using variable $VICBAR4A$. This variable captures direct victimization by criminal groups. The operationalization is as follows:

\begin{spacing}{1}
\begin{equation}
\text{Gang Victim} =
\begin{cases}
1, & \text{if respondent/family was a victim of extortion in last 12 months} \\
0, & \text{otherwise}
\end{cases}
\end{equation}
\end{spacing}

\vspace{1em}

\noindent Independent Variable 2 ($X_{State}$): Measures institutional abuse as exposure to state-led "kickback" (Bribery) using variable EXC6. This variable captures direct institutional corruption:

\begin{spacing}{1}
\begin{equation}
\text{Bribe Victim} =
\begin{cases}
1, & \text{if a public official requested a bribe in last 12 months} \\
0, & \text{otherwise}
\end{cases}
\end{equation}
\end{spacing}

\vspace{1em}

\noindent Moderator ($Z$): Economic Perception - Measured as the respondent's retrospective evaluation of their personal economic situation (IDIO2), serving as a proxy for the receipt of effective state investment. Coded on a 3-point ordinal scale:

\begin{spacing}{1}
\begin{equation}
\text{Econ Perception} =
\begin{cases}
1, & \text{Worse} \\
2, & \text{Same} \\
3, & \text{Better}
\end{cases}
\end{equation}
\end{spacing}

\vspace{1em}

\noindent Control Variables (C): To mitigate confounding, the following structural and political variables are included:

\noindent Education ($Inc_{proxy}$): A proxy for Socio-Economic Status (SES). Coded as a 4-point ordinal scale based on the highest level of education completed (edr):

\begin{spacing}{1}
\begin{equation}
\text{Education Level} =
\begin{cases}
0, & \text{None}, \\
1, & \text{Primary}, \\
2, & \text{Secondary}, \\
3, & \text{Higher Education}
\\
\end{cases}
\end{equation}
\end{spacing}

\vspace{1em}

\noindent Urbanization ($U$): Binary indicator for urban residence (ur1new), accounting for the concentration of both extortion activity and economic opportunity.

\begin{spacing}{1}
\begin{equation}
\text{Urban} =
\begin{cases}
1, & \text{if residence is Urban (City/Outskirts)} \\
0, & \text{if residence is Rural}
\\
\end{cases}
\end{equation}
\end{spacing}

\vspace{1em}

\noindent Corruption Perception ($C$): A generalized measure of how common the respondent believes corruption is among politicians (exc7). Reversed so higher values indicate higher cynicism:

\begin{spacing}{1}
\begin{equation}
\text{High Corruption Perc.} =
\begin{cases}
1, & \text{Very Uncommon} \\
2, & \text{Uncommon} \\
3, & \text{Common} \\
4, & \text{Very Common}
\end{cases}
\end{equation}
\end{spacing}

\vspace{1em}

\noindent Municipal Services ($W_{proxy}$): Satisfaction with local services (sgl1), serving as a proxy for the tangible delivery of public goods by the national government. Inverted so higher values indicate better services:

\begin{spacing}{1}
\begin{equation}
\text{Muni Services} =
\begin{cases}
1, & \text{Very Bad} \\
2, & \text{Bad} \\
3, & \text{Fair} \\
4, & \text{Good} \\
5, & \text{Very Good}
\end{cases}
\end{equation}
\end{spacing}

## Results

Results will be written up soon. In short I find that because the data is from 2021, it is hard to say much about democratic backsliding / transition (since the data is after the 2019 election of Bukele). That said I find those who are exposed to gang violence blame Bukele suggesting that voters point to the current ruling party despite their claims of success.

### Model Estimation

```{r}
#| output: asis
#| message: false

#Run logistic regression for H1
model_base <- glm(bukele_support ~ 
                    extortion_victim +             # IV
                    econ_perception +              # Moderator 
                    is_urban + 
                    edr + 
                    corruption_perc + 
                    muni_services,
                  data = LAPOP_Model, 
                  family = "binomial")

# 2. Run the Interaction Model (Testing H1)
# This adds the "Dampening" test
model_interaction <- glm(bukele_support ~ 
                           extortion_victim + 
                           econ_perception +
                           is_urban + 
                           edr + 
                           corruption_perc + 
                           muni_services +
                           extortion_victim * econ_perception ,
                         data = LAPOP_Model, 
                         family = "binomial")

# 3. Create the Side-by-Side Table
stargazer(model_base, model_interaction, 
          type = "latex",
          title = "Logistic Regression Results: Determinants of Bukele Support",
          column.labels = c("Main Effects", "Interaction (H3)"),
          
          # Labels for your variables
          covariate.labels = c("Extortion Victim", 
                               "Econ Perception (Better)", 
                               "Urban (Yes)", 
                               "Education (0-3)", 
                               "Corruption Perc (High)", 
                               "Muni Services (Better)",
                               "Interaction: Victim x Econ"),
          
          dep.var.labels = "Support for Bukele",
          star.cutoffs = c(0.05, 0.01, 0.001))

```

```{r}
#| output: asis

# HA3
model_bribe_base <- glm(bukele_support ~ 
                          bribe_victim + 
                          econ_perception + 
                          is_urban + 
                          edr + 
                          corruption_perc + 
                          muni_services,
                          data = LAPOP_Model, family = "binomial",
                        )

model_bribe_interaction <- glm(bukele_support ~ 
                                 bribe_victim * econ_perception + 
                                 is_urban + 
                                 edr + 
                                 corruption_perc + 
                                 muni_services,
                               data = LAPOP_Model, family = "binomial")

# 3. View Results
stargazer(model_bribe_base, model_bribe_interaction, type = "latex",
          covariate.labels = c("Bribe Victim (State)", "Econ Perception (Better)", 
                               "Urban", "Education", "Gen. Corruption Perc", "Muni Services",
                               "Interaction: Bribe x Econ"),
          dep.var.labels = "Support for Bukele")
```

\newpage

# Limitations

Note: This submission focuses on the LAPOP survey data from 2021. That said, there are a couple of limitations with this data set that need to be addressed before the finalization of this paper.

Currently, I am examining different approaches to aggregation and measures of my variables using the following data sets:

2021 Legislative Election Data (n = 262)

-   Municipality Vote share for each party
-   Bukele’s Party as opposed to direct vote

DIGESTYC - Census (Similar to ACS) 2021 - Municipality level demographic data

ACLED Conflict Data - Gang and State Violence events from 2018 to 2024

CSES 2019 Post-Election Survey (El Salvador) - Comparative Study of Electoral Systems (CSES)

LAPOP 2023 Survey Data (which has questions about the 2019 election which is far more interesting and relevant but seems to be biased)

\newpage

## Works Cited
