---
title: "Democratic Backsliding in El Salvador"
subtitle: "POL 194H: Honors Thesis"
author: "Luca Alioto"

format:
  pdf:
    documentclass: article
    pdf-engine: pdflatex
    fontsize: 12pt
    geometry:
      - margin=1in
    include-in-header:
      text: |
        \usepackage{tikz}
        \usetikzlibrary{calc, arrows.meta, positioning, shapes}
        \usepackage{caption}
        \usepackage{fancyhdr}
        \usepackage{titling}
        \usepackage{mathptmx}
        \usepackage{float}
        
        \setlength{\parindent}{2em}   % Controls indent size (standard is ~1.5–2em)
        % Hanno didnt use so in wokring paper so ig I wont as well
        % \usepackage{indentfirst}     % Also indent first paragraph after headings
        
        % --- REQUIRED FOR MANUALLY FIXING SPACING ---
        \usepackage{setspace} 
        \doublespacing       % Starts the document in double spacing
        % Fix double-indented lists
        \usepackage{enumitem}
        \setlist[itemize]{leftmargin=*}
        \setlist[enumerate]{leftmargin=*}
        % --------------------------------------------

        % ---- Fancy header setup ---- 
        \fancypagestyle{plain}[fancy]{}
        \fancyhf{}
        \fancyhead[L]{Alioto | POL 194HA}
        \fancyhead[C]{}
        \fancyhead[R]{\thepage}
        \renewcommand{\headrulewidth}{0.4pt}

        % ---- Center the Title/Author nicely ----
        \pretitle{\begin{center}\LARGE}
        \posttitle{\par\end{center}\vskip 0.5em}
        \preauthor{\begin{center}\large}
        \postauthor{\par\end{center}\vskip 1em}

execute:
  echo: false
  warning: false
  message: false

bibliography: references.bib
---

```{r setup, include=FALSE}
library(knitr)
library(here)
## opts_knit$set(root.dir = here::here("01_thesis_replication"))
```

```{r, include=FALSE}
# Load necessary libraries
library(tidyverse)
library(dplyr)
library(labelled)
library(kableExtra)
library(sjlabelled)
library(stringr)
library(haven)
library(readxl)
library(stargazer)
library(broom)
library(gt)

LAPOP <- haven::read_dta(here("01_data/02_lapop/SLV_2021_LAPOP.dta")) # Public Opinion data from 2021

LAPOP_A = LAPOP %>%
  filter(core_a_core_b == "Core A") # Group A respondents
```

```{r}

# Helper Function to remove LAPOP "Junk" codes
clean_lapop <- function(x) {
  case_when(
    x %in% c(888888, 988888, 999999) ~ NA_real_,
    TRUE ~ as.numeric(x)
  )
}

LAPOP_Model <- LAPOP_A %>%
  select(
    m1,       # B: DV - Bukele Approval
    vicbar4a, # Ex: IV - Extortion
    idio2,    # Ec: Moderator - Econ Perception
    edr,       # Inc (Proxy): Education Years
    ur1new,   # U: Confounder - Urban
    exc7,     # C: Control - Corruption Perception
    exc6,
    sgl1      # W (Proxy): Municipal Services
  ) %>%
  
 
  mutate(across(everything(), clean_lapop)) %>%
  
  # Step B: Create Analysis Variables
  mutate(
    # --- DV: Bukele Support (B) ---
    # 1=Support (Very Good/Good), 0=No Support (Regular/Bad/Very Bad)
    bukele_support = case_when(
      m1 %in% c(1, 2) ~ 1,
      m1 %in% c(3, 4, 5) ~ 0,
      TRUE ~ NA_real_
    ),
    
    # --- IV: Extortion Victim (Ex) ---
    # Original: 1=Yes, 2=No -> Recode to 1=Yes, 0=No
    extortion_victim = case_when(
      vicbar4a == 1 ~ 1,  # Yes
      vicbar4a == 0 ~ 0,  # No (The Codebook Value)
      TRUE ~ NA_real_
    ),
    
    # --- Moderator: Econ Perception (Ec) ---
    # Original: 1=Better, 2=Same, 3=Worse
    # REVERSE CODE: 3=Better, 2=Same, 1=Worse 
    # (Higher number = Better economy, making the interaction intuitive)
    econ_perception = case_when(
      idio2 == 1 ~ 3,
      idio2 == 2 ~ 2,
      idio2 == 3 ~ 1,
      TRUE ~ NA_real_
    ),
    
    # --- Proxy for Income (Inc): Education ---
    # Continuous variable (Years of schooling)
    
    # --- Confounder: Urban (U) ---
    # 1/2 = Urban, 3/4 = Rural -> Recode to 1=Urban, 0=Rural
    is_urban = case_when(
      ur1new %in% c(1, 2) ~ 1,
      ur1new %in% c(3, 4) ~ 0,
      TRUE ~ NA_real_
    ),
    
    # --- Control: Corruption Perception (C) ---
    # Original: 1=Very Common ... 4=Uncommon
    # REVERSE CODE: 4=High Corruption ... 1=Low Corruption
    corruption_perc = case_when(
      exc7 == 1 ~ 4,
      exc7 == 2 ~ 3,
      exc7 == 3 ~ 2,
      exc7 == 4 ~ 1,
      TRUE ~ NA_real_
    ),
    
    edr = edr, # 0 as none 3 as Uni 
    bribe_victim = exc6, 
    
    # --- Proxy for Welfare (W): Muni Services ---
    muni_services = case_when(
      sgl1 == 1 ~ 5,
      sgl1 == 2 ~ 4,
      sgl1 == 3 ~ 3,
      sgl1 == 4 ~ 2,
      sgl1 == 5 ~ 1,
      TRUE ~ NA_real_
    ),
  )

```

## Introduction

Some believe that the canonizing event in the democratic backsliding of El Salvador began on March 26th, 2022, a day that saw the highest mortality rate in two decades. In the span of 24 hours, sixty-two people were killed in a gang attacks; a violent breakdown in government-gang negotiations [@gellman2022]. 81st President Nayib Bukele responded by declaring a “war on gangs,” establishing a State of Exception that suspended the constitutional rights to freedom of assembly, privacy in communications, and due process. In the weeks that followed, military officials were deployed in the streets, and authorities arrested over 8,500 people—a number that would increase to nearly 80,000, or more than 1% of the population, by 2024. As Bukele's response caused widespread international condemnation and concerns over arbitrary detention, his domestic approval rating soared to nearly 90%, showing that his iron fist approach to gangs and gang violence was backed by the citizens of El Salvador [@gellman2022; @meléndez-sánchez2024].

While this marks an escalation in Bukele's centralization of power this was merely the dramatic display of the erosion. Research from scholars and international observers suggests that it began much earlier, when his administration worked to systematically undermine opposition parties and lower the efficacy of courts, independent agencies, local governments, civil society organization, and the press. Scholars argue the country’s authoritarian shift can be marked by the rapid acceleration in the centralization of power, occurring after his party, Nuevas Ideas, secured a legislative supermajority in 2021 [@meléndez-sánchez2022]. Following the 2021 legislative elections, the Assembly swiftly removed all five magistrates of the Constitutional Chamber of the Supreme Court and the Attorney General, replacing them with loyalists [@u.s.departmentofstateHumanRightsReports2021; @gellman2022]. Later that year, the reconstituted court authorized presidential reelection despite explicit constitutional prohibitions, paving the way for Bukele’s successful 2024 bid which resulted in his re-introduction to a government that had been radically reshaped [@bbc2024].

Long before Bukele entered office, Salvadorans had grown deeply disillusioned with representative democracy. In 2018, the year before his election, 63.4 percent reported dissatisfaction with democracy [@lapoplabAmericasBarometerSalvador2018]. More than 60 percent agreed that elections were “a waste of time,” and nearly 80 percent said major parties did not represent people like them [@iudop2018; @meléndez-sánchez2022]. Support for traditional parties had collapsed: just 30.8 percent identified with two of the promenient political parites by 2018. Public distrust extended further, as 84.9 percent of Salvadorans believed that at least half of all politicians were corrupt [@lapoplabAmericasBarometerSalvador2018]. In addition to perceptions of widespread corruption, Salvadorans faced large levels of insecurity and exposure to violence. Between 2004 and 2018, crime and violence consistently emerged as the country’s most severe problem, rising from 32 to 62 percent in national opinion surveys. Even during periods of declining homicide rates—such as the 2012–2014 truce—between 15 and 20 percent of Salvadorans reported living in neighborhoods controlled by gangs [@lapoplabAmericasBarometerSalvador2018; @castro].

Previous governments failed to contain violence and corruption, producing a widening gap between Salvadorans and the country’s democratic institutions. Bukele filled that void through appealing to widespread frustration and insecurity [@vanderborgh2019; @humanrightswatch2023]. He won 53 percent of the vote—21 points ahead of his nearest competition [@márquez2021]. His enduring popularity had gone beyond the campaign trail, steming not merely from rhetoric but from his ability to deliver on it. Although the March 2022 security crackdown generated significant human-rights concerns, it achieved what previous administrations were unable to. By early 2023 the gangs that had long dominated territory, extorted residents, and driven El Salvador’s status as one of the world’s most violent countries had been largely dismantled. Official homicide and extortion rates fell to historic lows. While Bukele is not the first Latin American president to adopt iron fist policies, he is the first to do so with outcomes that, from the public’s perspective, appeared to decisively resolve the very insecurities that had eroded trust in the democratic system [@meléndez-sánchez2022].

Bukele has roooted his political project in the claim that he, not existing institutions, embodies the true will of the Salvadoran people. He capitalized on widespread anger toward corruption with slogans such as “Return what you have stolen” and “There is enough money when no one steals". In speeches, he routinely frames his administration as a historic effort to restore “real democracy” after two centuries of elite manipulation and institutional decay. “Now we are building a real democracy,” he proclaimed. “For 200 years, democracy was a pantomime… They never cared about people, they only cared about votes” [@meléndez-sánchez2022]. Following his removal of term limits and subsequent re-election, Bukele went further, asserting that democratic legitimacy derived exclusively from his overwhelming popular mandate noting,

> “This is the first time that a single party exists in a country in a fully democratic system. The entire opposition was pulverized… Democracy means the power of the people, demos and kratos. We Salvadorans are united,”

\noindent labeling his consolidation of power and removal of both of the longstanding major parties from participation as democratic [@meléndez-sánchez2022; @wolf2024]

Largely, public opinion trends suggest that many Salvadorans have come to accept this reframing. Between 2018 and 2021, satisfaction with democracy rose sharply, and distrust in elections declined. In effect, Salvadorans have reported growing confidence in what they perceive to be a democratic system at the very moment that core democratic norms and institutions are being weakened[@lapoplabAmericasBarometerSalvador2018; @meléndez-sánchez2023].

As perceptions of democracy concentrate around the executive, and his approach to creating "the safest country in North America," understanding the mechanisms sustaining such high support is critical [@velez2024]. This raises a fundamental question for the study of democratic backsliding in contexts of high violence:

> **How does exposure to state and criminal violence influence support for President Bukele and authoritarian policies in El Salvador?**

\noindent This study examines whether a dynamic of performance legitimacy and selective accountability sustains Bukele's popularity during this period of democratic erosion. By analyzing individual-level data from the 2021 AmericasBarometer (LAPOP), I explore the association between exposure to violence—both criminal and state-led—and support for the incumbent. As well, I work to understand the effect economic perceptions and the expansion of Bukele's welfare state act as compensatory mechanisms, potentially overriding the political costs of security failures.

### Research Design

To answer this question, I utilize a comparative Logistic Regression design on a national probability sample of Salvadoran adults ($N=1,435$). I operationalize victimization in two distinct ways—Gang Extortion (representing policy failure) and State Bribery (representing institutional abuse)—to test if voters attribute blame differently based on the perpetrator. To isolate these mechanisms, I control for structural confounders including urbanization and education (serving as a proxy for socioeconomic status) and test for the moderating role of retrospective economic perception.

### Key Findings

In this paper, I find that victims crime do not blindly demand authoritarianism or call for iron fist policies. Instead, my results reveal a process of selective accountability. I find that victims of gang extortion are significantly less likely to support President Bukele ($\beta = -0.465$), treating continued insecurity as a specific policy failure. However, voters do not penalize the administration for state-led corruption, suggesting Bukele is largley insulated from the reputational costs of institutional abuse. Ultimately, I find that economic performance serves as a critical buffer, with positive economic perceptions offsetting the disapproval generated by security failures, effectively buying the political capital necessary to weather the costs of violence.

## Literature Review

### Civil War, the Peace Accords, and the Long Term Institutional Effects

While many foreign observers initially regarded the Salvadoran Peace Accords of 1992 as Latin America’s most successful democratic transitions, scholars have since identified the pact as a source of institutional instability and public disillusionment with democracy[@buiza2018; @meléndez-sánchez2021]. These negotiations preceded a brutal civil war between the insurgent Farabundo Martí National Liberation Front (FMLN) and the Salvadoran government, administered in its final years by the right-wing Nationalist Republican Alliance (ARENA). Throughout the twelve-year period, the conflict claimed the lives of 75,000 civilian and thousands of combatants alongside displacing forcefully displaced nearly one million people [@chávez2015].

The civil war itself was rooted in decades of extreme socioeconomic inequality, elite domination, and systematic state repression under what critics label authoritarian military-oligarchic regime [@chávez2015]. In this context, the FMLN—a coalition of leftist guerrilla organizations—emerged with the objective of overthrowing the authoritarian state and fundamentally redistribute political and economic power. The founding of the FMLN in 1980 represented a pivotal convergence of the “Old Left” and the “New Left,” creating a group with substantial organizational capacity, military and political expertise, and international networks. While this ideological and strategic diversity was the groups initial strength, throughout the conflict leaders struggled to maintain unified positions on revolutionary strategy, negotiation, and long-term political goals. At the same time, the ideological decline of socialist parties in Latin America and Europe pushed the FMLN toward a more pragmatic orientation and away from its origional rigid Marxist-Leninist principles [@chávez2015; @meléndez-sánchez2021].

This consestions largely influenced the peace negotiations in 1992, where the organization set aside its earlier insistence on structural socioeconomic transformation. Instead, FMLN leadership articulated two core negotiating aims: the comprehensive demilitarization of the Salvadoran state and the creation of institutional guarantees for a competitive democratic order. The UN-mediated negotiation between the government of President Alfredo Cristiani and the FMLN ended the conflict in 1992, paving the way for the only sustained democratic period in Salvadoran [@chávez2015].

In their paper, @meléndez-sánchez2022 highlights three central components of the transition pact that initially enabled the emergence of a strong and stable party system following the peace accords. The first was the peace accords themselves, which transformed the FMLN from a mobilized guerrilla force into a political party able to participate freely in the political sphere. Second, the 1993 General Amnesty Law granted combatants on both sides immunity from prosecution for war-related crimes, allowing the leadership to remain in power after the war. Finally, a new Electoral Code, written in 1992, established high barriers to entry for new parties and gave leaders of both ARENA and the FMLN significant influence over future electoral processes.

Scholarship argues that these “pacted transitions,” aimed at safeguarding the interests of powerful elites, are key to stabilizing new democracies and creating “a sufficiently strong consensus about the rules of the game” so that no major elite actor is incentivized to revert to authoritarianism [@karl1990; @meléndez-sánchez2022]. This approach to building state capacity rests on the idea that conflict is rooted in weak institutional frameworks, leading external actors to focus their efforts on strengthening institutional capacity through state-building and democracy-support measures [@fox2012; @zulueta-fülscher2018]. The three components of El Salvador’s transition followed this logic, successfully persuaded the parties to lay down their arms, created conditions “minimally safe for powerful actors on both sides of the regime divide,” and enabled two historical rivals to form a democratic system [@meléndez-sánchez2022].

Nonetheless, scholarship suggests that when capacity-building becomes the primary focus of development plans, elites often sideline efforts to make institutional frameworks more inclusive or democratic. Increaing representation, participation, openness, and accountability would require structural reforms that depend on elites relinquishing at least part of their power to either the public or emerging political actors [@zulueta-fülscher2018]. More broadly,

> *“there is no necessity for a society to develop or adopt the institutions that are best for economic growth or the welfare of its citizens, because other institutions may be even better for those who control politics and political institutions”* [@daronacemoglu2013].

\noindent with being elites likely to resist and reforms which reduce their institutional control [@zulueta-fülscher2018].

Over time, the high barriers to entry and protections for existing leadership built into the 1992 peace accords encouraged a form of democratic consolidation in which a small group of elites on both sides of the regime divide could reliably safeguard their interests through ARENA and the FMLN. A key factor was the Amnesty Law, which allowed wartime elites to remain in control of both parties and prevented perpetrators of war crimes and corruption from being held accountable. Additionally, the Electoral Code gave party leaders full control over internal nomination and discipline, enabling these elites to remain in power for decades while removing any emerging opposition [@meléndez-sánchez2022]. This control extended across the broader political system. @gutiérrezsalazar2015 notes that postwar reforms did little to limit partisan influence over the judiciary, pointing to limited autonomy and extensive clientelist networks as key impediments to the maintenance of democratic norms.

These arrangements also preserved and deepened longstanding socioeconomic and political inequalities, allowing various forms of elite domination to persist and evolve in the following decades. As a result, inequality, economic exclusion, and social disparities remained entrenched, driving increases in external economic institutions through migration and and a large dependence on remittances .

while enduring economic exclusion and weak horizontal accountability persisted alongside it, creating conditions for long-term instability.

### Visible Corruption: Democratic Consolidation and Narrow Consensus

Democratic consolidation also produced a narrow procedural consensus, marked by limited ideological diversity between parties. In their papers, @colalongo2024 and @meléndez-sánchez2023 point to the 2009 election as a key example of this dynamic. After twenty years of ARENA governments, the FMLN won the presidency for the first time since the democratic transition and formed a left-wing administration under Mauricio Funes, a political outsider to the party. Funes’s election was widely interpreted as a compromise, signaling that the FMLN had sacrificed ideological purity to win the presidency for the first time. In hopes of curtailing this and retaining internal power, party leadership had initially vetoed Funes' nomination, later permitting him to run only on the condition that his running mate be a former FMLN civil war commander.

Despite his reformist platform, Funes largely replicated the political model of his ARENA predecessors, and party brands began to dilute. The FMLN, whose identity had long been built around opposing ARENA’s economic and security policies, ultimately favored policy continuity over reform. Contrary to campaign promises, the party did not attempt to de-dollarize the economy, nationalize key industries, or intensify redistribution; instead, Funes instituted a government austerity plan in his first month in office [@meléndez-sánchez2023]. Economically, ongoing market liberalization continued to erode state institutions, sustaining structural poverty and contributing to social violence [@colalongo2024].

This dilution extended to security policy. After experimenting with alternative approaches, Funes ultimately embraced the hardline gang crackdowns of previous ARENA governments. Funes’s decision to militarize public security in response to renewed gang offensives further entrenched this overlap, paving the way for deeper institutional violence [@roque2021; @colalongo2024]. This was only intensified after Funes was succeeded by Sánchez Cerén, who, like his predecessor, continued to favor free-market policies and repressive security approaches [@meléndez-sánchez2023].

In this environment, civic–military alliances persisted to preserve the status quo: elites protected their interests, constrained reforms, and prevented substantive debate over the country’s economic model. @chávez2015 notes that the FMLN’s turn toward pragmatism—and the concessions made to consolidate itself as a legal political party—anchored it firmly within ARENA’s neoliberal framework rather than its original transformative project. The presidency of Mauricio Funes thus marked an institutional shift away from the FMLN’s socio-economic reform agenda, solidifying its role as a center-left party that had conceded the economic status quo.

Alongside brand dilution, corruption investigations involving members of both ARENA and the FMLN reinforced voters’ perceptions that the two parties had become increasingly indistinguishable in their policy orientations and political behavior. A major catalyst for this disillusionment was a string of high-profile corruption cases beginning in 2013.

In September of that year, former President Francisco Flores (1999–2004) was accused of redirecting \$15 million in international earthquake-relief donations to ARENA’s campaign coffers. What followed was unprecedented: within a five-year span, six of the country’s most powerful political figures were publicly and credibly implicated in corruption. While in theory, high-profile corruption prosecutions can strengthen democracy by demonstrating institutional accountability and deterring future abuses, they can also backfire by generating a broader perception that the entire political system is corrupt. Scholars describe this as a sense that “the whole system is rotten,” a dynamic which proved especially destabilizing, as it coincided with the collapse of meaningful programmatic differences between ARENA and the FMLN. The result was a voter base increasingly disillusioned with the existing democratic structure which seemed aligned in their ideological and material goals [@meléndez-sánchez2022] .

These instances of corruption and alignment were not merely a matter of policy missteps but reflected a deeper structural feature of the postwar regime, one fundamentally centered on maintaining order rather than expanding representation or accountability. El Salvador’s democracy became, in effect, “hollowed out” by design— providing populist candidates little push back when attacking a democracy that failed to move beyond capacity-building.

### Gangs, Mano Duro, and the Future for Bukele 

The gangs - Who are they

The majority of gang members are male, unemployed, live in urban areas, and have dropped out of school (Cruz, 2005; Olate et al., 2011). Though El Salvador’s two primary gangs are portrayed by the mainstream media as the main actors contributing to high levels of crime and delinquency, gangs are estimated to have accounted for less than a third of the homicides and extortions in El Salvador in 2009

Gangs are primarily involved in small-scale localized crime and delinquency; there is no evidence of their involvement in large-scale transnational drug trafficking (UNODC, 2010). Examples of their illegal activities include: use of illegal drugs and small-scale trafficking, petty theft, muggings, extortion of local businesses and drivers of public transportation vehicles, control of territories, use of illegal weaponry, and murders, particularly of members of the rival gang (UNODC, 2007).

Iron Fist (the past)

@olate2012

The Salvadoran government has launched several anti-gang policies also known as Mano Dura (Iron Fist) (2003), Super Mano Dura (2004), and recently the Proscripción de Pandillas (2010). One of the direct consequences of the implementation of these policies has been the steady increase in the incarcerated population, especially gang members (USAID, 2006). Owing to limitations in data collection, underreporting, and political factors, determining an accurate number of active gang members is an elusive task; however, official statistics suggest that there are approximately 8500 gang members (Policía Nacional Civil, 2010) in prison in El Salvador,2 most from MS and 18. The number of youth involved in gangs outside of prison remains uncertain. Although Iron Fist policies have been widely supported by political actors and the Salvadoran population, their results are less than optimal (PNUD, 2009). The judiciary and prison systems do not have the resources and institutional capacity to face the high demand produced by the implementation of these policies. Prison conditions, especially for those exclusively receiving gang members, have reached unsanitary levels due to overcrowding (Seelke, 2011). In addition, the level of crime decreased only temporarily and, due to the level of repression, the gangs have become more organized and violent (Aguilar, 2006). @olate2012

The phenomenon of "mano dura" politics, as detailed by Wolf (2017) and Cruz & Durán-Martínez (2016), has deep roots in El Salvador and the region. Mass incarceration and criminalization of marginalized communities have been common state responses to gang violence, often producing short-term calm but long-term democratic deterioration. Scholars have also explored how pacts between the state and criminal groups, like the 2012 gang truce, conceal rather than resolve violence. Cruz (2023) highlights the authoritarian drift in Central America, showing how tough-on-crime policies have historically empowered the military and police at the expense of human rights. As well, scholars of penal populism (e.g., Rosen et al., 2022) show that fear of crime erodes support for liberal democratic norms, making repressive measures popular among voters traumatized by violence.

Successive governments across the political spectrum pursued a mano dura (“iron fist”) approach to gangs and crime, typically with the intent to gain an electoral advantage. Combining joint police and military patrols with area sweeps and mass arrests, the strategy contributed to a rise in violence and transformations of the gangs (Wolf 2017). The first FMLN administration was prepared to discontinue gang suppression yet unable to make headway in social prevention. In 2012 it began secret negotiations with the gangs to achieve a reduction in homicides in exchange for improved prison conditions and other benefits. When the truce collapsed after a year, murders escalated to an unprecedented rate of 106 per 100,000 inhabitants in 2015

Bukele  - what is gonna happen

Short lived success?

Based on visits to former gang strongholds in 10 municipalities, analysts found that emergency rule has disrupted the groups’ leadership structures and demobilized much of their street-level membership. However, dozens of gang cliques remain active and extortion continues, albeit on a smaller scale (InSight Crime 2023). In previously gang-dominated communities there is a clear sense of respite, and opinion polls reveal overwhelming support for the state of emergency (IUDOP 2024a).

A need to keep the state of emergency going (public opinion)

But Security Minister Gustavo Villatoro warned that the measure would continue until the gang threat had been completely eradicated (Urbina 2024). The official’s remarks point to an electoral autocracy’s core dilemma: the risk of becoming obsolete when it fails or when it succeeds in resolving the crisis that justifies its existence. To maintain popular consent, an autocratic regime must maintain yet also appear to end the crisis that legitimizes its rule (Matovski 2021). Bukele wants to attribute his security achievements to the state of emergency. At the same time, he raises the specter of gang violence to justify extending the measure that gives him widespread support, with Salvadorans giving him a rating of 8.5 on a scale of 0 to 10 (IUDOP 2024a: 35).

And what happens if it fails

While Bukele publicly condemned his predecessor’s policy choices, in private he embraced them. As investigations by the independent news outlet El Faro revealed, his administration made an undisclosed pact with gang leaders to obtain electoral support and a drop in violence (Martínez, Cáceres, and Martínez 2021). The mass killings in late March 2022 were a retaliation by gangs for the breakdown of the agreement after the authorities had arrested gang members heading to Guatemala in an official vehicle (Martínez 2022).

Outside of this, its not going that great...

When he ran for his first presidential term, Bukele vowed to build 15 megaprojects at an estimated cost of at least \$1,525 million (Labrador 2023a). It appears he has since postponed if not shelved many of his ambitious plans, including those for transport facilities, schools and university campuses, and hospitals (Peñate 2024). The CECOT, by contrast, was not a campaign promise but features in the publicity surrounding the state of emergency. It was apparently deemed more suited to portraying Bukele as a capable leader.

The authorities have since arrested more than 77,000 people (Valencia 2024), but the lack of publicly available statistics makes it difficult to confirm possible gang affiliations or the age and gender of detainees (Viñas and Budasoff 2024d)

Corruption

Bukele had vowed to personally put in prison anyone who touched a cent. Yet corruption, and impunity for wrongdoing, have become hallmarks of the regime. Its practices, some old and some new, include the propagandistic use of uncompleted public works, opaque government spending, and contracts that benefit relatives and Nuevas Ideas members (Redacción de El Faro 2023).

While some infrastructure projects serve his political project, the history of others shows how Bukele uses his political power for personal financial gain. One of these is Surf City, a key tourism development initiative at El Salvador’s Pacific Coast.

For the new ruling group, land speculation and property development are a means to gain capital, even when these deals have a detrimental impact on the ecosystem and water supplies, as in the case of Surf City. The ministers of the environment, public works, and housing appear to have been appointed not for their expertise but for their loyalty to President Bukele and their willingness to sidestep regulations (Gutiérrez 2024).

Is he cooked? \
Almost half the population experiences food insecurity, and between 2019 and 2022, extreme poverty increased by two percentage points to 8.7% (Ribando Seelke 2024). The economy depends heavily on remittances (24% of GDP) and in 2023 recorded a growth of only 2.6%, making it the least dynamic in Central America (González Díaz 2024)

Surveys indicate that Salvadorans are now more worried about the lack of jobs and the rising cost of living than about security matters. In 2023, 70% of respondents to the Jesuit University’s end-of-year poll identified economic issues as the nation’s main problem (IUDOP 2024a).

The reduction in crime and violence associated with the state of emergency has prompted greater domestic and diaspora investment, but foreign direct investment continues to remain low (Ribando Seelke 2024). Meanwhile, government debt exceeds 90% of GDP, and it is unclear how the country will make its debt repayments in the coming years (González Díaz 2024

### Gaps in the Literature: Public Opinion and Democratic Perceptions

Despite growing literature, a gap remains in understanding how public opinion responds to state violence and how citizen tolerance for authoritarianism is shaped by their experiences and identities. Most studies focus on policy effects or elite behavior, with fewer examining how support for authoritarian regimes is built and maintained from below. In their thesis, Castro and Kotti (2022) examine the relationship between Gangs and Electoral participation, arguing that gang presence in a community has a large effect on electoral outcomes.

Consequently, acts of democratic erosion have had little negative impact on public opinion. Public satisfaction with democracy has actually increased since his election. Between 2018 and 2021, the proportion of Salvadorans reporting satisfaction with democracy rose sharply, while mistrust in elections declined. In effect, Salvadorans have grown more confident in a political system that is, by objective institutional standards, becoming less democratic.

Evidence suggests that Salvadorans’ attitudes toward democracy are increasingly shaped not by institutional performance or policy outcomes, but by their evaluations of a single political actor: the executive. Positive perceptions of Bukele—more than trust in institutions or improvements in state capacity—strongly predict satisfaction with democracy and greater tolerance for executive concentration of power. Far from signaling a renewed belief in democratic governance, rising public approval appears to reflect an increasingly personalized understanding of political legitimacy—one in which the president, rather than democratic institutions, anchors citizens’ perceptions of the democracy.

Much said about democracy but not about bukele himself.

As well, there is a large gap in research on the effects of state violence in El Salvador on voting behavior and the long-term effects of Bukele’s mass arrests on his popularity. Not much literature is capturing if public opinion is shaped soley by success, and if failure in providing safety and security will cause voters to look beyond rhetoric. If he is suceptible to the same failures as ARENA and FMLN,

## Causal Pathways & Theoretical Motivation

To examine the association between exposure to violence and economic conditions and support for democratic governance under Bukele, I employ a quantitative research design utilizing survey data from the 2021 LAPOP AmericasBarometer for El Salvador. The analysis focuses on individual-level responses to questions regarding experiences of state violence, perceptions of economic well-being, and attitudes toward democracy.

### Theory: Data-Generating Process (DGP)

To understand the determinants of incumbent support in the context of democratic backsliding, I propose a Data-Generating Process (DGP) where individual political preference ($Y$) is a function of exposure to violence ($X$), conditional on the voter's economic satisfaction ($Z$).

\noindent The causal mechanism linking violence to support for Bukele is conceptualized as follows:

\noindent \*\*Exposure to Violence\*\* ($X$): Exposure to insecurity serves as the primary independent variable. I theorize that the source of violence determines the political reaction:

-   Gang Extortion ($X_{Gang}$): Following the "Iron Fist" literature, exposure to non-state violence is hypothesized to increase perceptions of insecurity, potentially driving a demand for authoritarian order and increasing support for the strongman ($H_1$).
-   State Predation ($X_{State}$): Conversely, exposure to state-led predation (Bribery) represents a violation of the social contract. This experience is hypothesized to decrease satisfaction with democratic governance and reduce incumbent support ($H_2$).

\noindent \*\*Economic Perceptions\*\* ($Z$): Economic satisfaction acts as a moderating variable. Individuals with positive economic perceptions are influenced by perceptions of the administration's management of the economy and distribution of economic success. I hypothesize that this "performance legitimacy" weakens the relationship between violence and political preference. Specifically, positive economic evaluations may neutralize the dissatisfaction caused by state violence or amplify the support derived from security demands ($H_3$).

\noindent \*\*Support for Bukele\*\* ($Y$): The outcome variable represents the individual's binary approval of Bukele’s governance. It is modeled as the net result of these competing pressures: the security experience ($X$) and the economic experience ($Z$).

\noindent \*\*Control Variables\*\* ($\mathbf{C}$):To block confounding paths and isolate the causal effect of violence, I condition on the following structural factors:

-   Urbanization: Urban residents may have distinct exposure levels to extortion networks and possess different baseline political attitudes compared to rural residents.
-   Income (SES): Socioeconomic status influences both the probability of victimization (target suitability) and political preferences. (Operationalized via Education).
-   Perceptions of Corruption: An individual's general view on political corruption is included to distinguish specific victimization events from generalized anti-system cynicism.

\vspace{1em}

These hypothesized relationships are visualized in the Directed Acyclic Graph (DAG) presented in Figure 1. The diagram illustrates the causal pathways linking Victimization ($V$) to Incumbent Support ($B$), while explicitly mapping the structural confounders—such as Urbanization ($U$) and Income ($Inc$)—that simultaneously influence the likelihood of victimization and political attitudes. The dashed line connecting Economic Perception ($Ec$) to the main pathway represents the theorized moderating effect ($H_3$).

\begin{center}
\begin{tikzpicture}[
    % Styles
    var/.style={circle, draw=none, fill=none, inner sep=2pt, font=\large\bfseries},
    direct/.style={->, >=Latex, thick},
    interact/.style={->, >=Latex, thick, dashed}
]

    % --- MAIN AXIS (Centered on 0,0) ---
    \node[var] (v0) at (-3, 0) {V};   % IV: Extortion/Violence
    \node[var] (v1) at (3, 0) {B};     % DV: Bukele Support (Right)

    % --- THE MODERATOR (Centered Top) ---
    \node[var] (v7) at (0, 1) {Ec};    % Ec: Econ Perceptions

    % --- LEFT SIDE (Background/Confounders) ---
    \node[var] (v2) at (-3, 1.5) {U};    % Urban (Above Ex)
    \node[var] (v4) at (-1.5, 2.5) {Inc}; % Income (Higher up, between U and Ec)
    \node[var] (v3) at (-1.5, -1.5) {W};  % Welfare (Below path)

    % --- RIGHT SIDE (Outcomes/Controls) ---
    % REPLACED Id WITH C HERE (Top Right)
    \node[var] (v5) at (3, 1.5) {C};     % Corruption Perception

    % --- ARROWS ---
    
    % 1. The Main Path (Ex -> B)
    \draw[direct] (v0) -- (v1) coordinate[midway] (midpoint);
    
    % 2. The Interaction (Ec -> Center)
    \draw[interact] (v7) -- (midpoint);
    
    % 3. Moderator Direct Effect
    \draw[direct] (v7) -- (v1);

    % --- CONTROL PATHS ---
    
    % Urban (The Confounder)
    \draw[direct] (v2) -- (v0);   % U -> Ex
    \draw[direct] (v2) -- (v7);   % U -> Ec
    \draw[direct] (v2) -- (v5);   % U -> C (Urban affects Corruption Perc)
    \draw[direct] (v2) -- (v4);   % U -> Inc
    
    % Income
    \draw[direct] (v4) -- (v7);   % Inc -> Ec
    \draw[direct] (v4) -- (v5);   % Inc -> C (Income affects Corruption Perc)
    
    % Corruption (Formerly Ideology) -> Bukele
    \draw[direct] (v5) -- (v1);
    
    % Welfare Path
    \draw[direct] (v0) -- (v3);   % Ex -> W
    \draw[direct] (v3) -- (v1);   % W -> B
    
\end{tikzpicture}
\end{center}
\begin{spacing}{1.25}
\begin{tikzpicture}
\node[align=left, font=\small, anchor=west] at (0,0) {
\textbf{Legend} \\
\textbf{V} = Exposure to Victimization (IV) \\
\textbf{B} = Support for Bukele (DV) \\
\textbf{Ec} = Economic Perceptions (Moderator) \\
U = Urbanization \\
Inc = Income \\
C = Perceptions of Corruption \\
W = Exposure to Welfare
};
\end{tikzpicture}
\end{spacing}

### Model Specification

\noindent To test these relationships, I estimate the probability of incumbent support using a Logistic Regression model. The log-odds of support are modeled as:

$$\ln \left( \frac{P(Y_i=1)}{1-P(Y_i=1)} \right) = \beta_0 + \beta_1 X_i + \beta_2 Z_i + \beta_3 (X_i \cdot Z_i) + \mathbf{\gamma C_i} + \epsilon_i$$

\noindent Where:

\begin{spacing}{1.25}
$Y_i$: Binary indicator of Support for Bukele.

$X_i$: Exposure to Violence (modeled separately as $X_{Gang}$ or $X_{State}$).

$Z_i$: Retrospective Economic Perception (Proxy for State Investment).

$X_i \cdot Z_i$: Interaction term testing the compensatory/dampening hypothesis.

$\mathbf{C_i}$: Vector of controls (Urbanization, Education, Corruption Perception, Municipal Services).
\end{spacing}

\noindent Building on this theoretical framework, I test the following hypotheses:

> $H_0$: There is no statistically significant relationship between exposure to victimization and the probability of supporting President Bukele.

#### The "Iron Fist" Hypothesis (Non-State Violence)

\noindent Literature on crime and authoritarianism suggests that high levels of criminal insecurity drive citizens to demand "Mano Dura" (Iron Fist) policies. Under conditions of high gang prevalence, victims are often willing to trade civil liberties and democratic checks for the promise of order. As a result, victimization by non-state actors is expected to increase the appeal of the strongman.

> $H_1$: Individuals exposed to non-state violence (Gang Extortion) will be more likely to support President Bukele, reflecting a demand for authoritarian security.

#### The "Broken Contract" Hypothesis (State Violence)

\noindent On the other hand, direct predication by state actors represents a violation of the social contract no matter the perceived benefits of violence. While citizens may tolerate strict measures against criminals, standard accountability theory suggests they will penalize the incumbent when the state apparatus itself becomes predatory. Therefore, victims of institutional corruption are expected to distinguish between the leader's rhetoric and the state's actual behavior.

> $H_2$: Individuals exposed to state-led predation (Bribery) will be less likely to support President Bukele, reflecting dissatisfaction with institutional corruption.

#### The "Performance Legitimacy" Hypothesis (Moderation)

\noindent Finally, I hypothesize that the political costs of security failures, most notably the disaproval caused by state violence, can be offset by economic performance. Following the logic of "performance legitimacy," perceived economic improvements serve as a compensatory mechanism. I hypothesize that economic satisfaction will moderate the relationship between victimization and support, effectively "dampening" the grievance caused by exposure to violence.

> $H_3$: Positive economic perceptions will weaken the relationship between exposure to victimization and support for President Bukele.

## Data and Measurement

### Unit of Analysis

The unit of analysis for this study is the individual survey respondent, representing the voting-age population (18 years and older) of El Salvador.

### Data Source and Coverage

To examine the small-scale foundations of authoritarian support, I rely on data from the 2021 AmericasBarometer conducted by the Latin American Public Opinion Project (LAPOP). This survey utilizes a stratified, multi-stage cluster probability design to ensure national representatives across both urban and rural strata. The fieldwork was conducted via face-to-face interviews, capturing public sentiment during the consolidation of the Bukele administration but prior to the implementation of the 2022 State of Exception.

### Sample Restrictions

The 2021 AmericasBarometer employs a "split-sample" design (Core A and Core B) to accommodate a broader range of questions. This study utilizes the Core A module, which contains the primary variables regarding victimization and political attitudes. While the total survey sample includes roughly 3,000 respondents, the Core A subset consists of approximately 1,500 individuals. After performing list-wise deletion for missing values on key theoretical variables—specifically victim status and economic perception—the final analytical sample consists of $N = 1,435$ observations.

As well, the analysis excludes responses coded as "Don't Know" $(888888)$, "No Answer" $(988888)$, and "Inapplicable" $(999999)$. These codes represent non-substantive responses that do not map onto the theoretical continua of interest (e.g., the ordinal scale of economic perception or the binary status of victimization). For the "Inapplicable" category, missingness is structural, resulting from survey skip patterns (e.g., questions asked only to specific subsets of respondents). To ensure model validity, these observations are treated as missing data (NA) and removed from the final analytical sample.

### Limitations

While the individual-level approach allows for a direct assessment of the mechanisms linking victimization to political support, the structure of the data limits the analysis to a single point in time. This is especially pertinent given the rapidly evolving political landscape in El Salvador under Bukele, and due to the potential for reverse causality between political attitudes and reported victimization.

As well, the data fails to capture the attitudes of respondents before the 2019 election, where the question of Bukele support would be most relevant. Data from the 2023 LAPOP survey includes a question on vote choice in 2019, but the responses were heavily skewed toward Bukele, with over 70% of respondents reporting voting for him despite only winning around 52% of the vote. This suggests significant social desirability bias in the retrospective vote choice question, and puts into question the validity of using both this retroactive reporting as well as perception variables as a whole as the DV.

Additionally, the split-sample design prevents the entire survey from being utilized, requiring the use of proxies back primarily by theories around the data generating process for certain demographic controls (e.g., Education for Income) that were located in the excluded survey module.

### Key Variables and Coding

This study focuses on the following key variables from the LAPOP 2021 Core A survey and are operationalized as followed:

\noindent Dependent Variable ($Y$): Support for Bukele - Measured as a binary indicator of job approval for President Bukele. Derived from variable $M1$ ("Job Approval") in the 2021 LAPOP survey. To provide a rigorous test of committed support, neutral responses are grouped with the opposition:

\begin{spacing}{1.25}
\begin{equation}
\text{Support for Bukele} =
\begin{cases}
1, & \text{if respondent rates performance as "Good" or "Very Good"} \\
0, & \text{if respondent rates performance as "Fair", "Bad", or "Very Bad"}
\end{cases}
\end{equation}
\end{spacing}

\vspace{1em}

The decision to group neutral responses ("Fair") with the opposition reflects two considerations specific to the Salvadoran context. Methodologically, given the "ceiling effect" of President Bukele's super-majoritarian approval ratings in 2021, treating neutral responses as support would artificially inflate the dependent variable, reducing variance and statistical leverage.

Theoretically, literature on survey behavior in backsliding democracies suggests that "neutral" categories often serve as a "safe harbor" for critical respondents subject to social desirability bias or fear of reprisal [@brownback; @tannenberg2017]. Therefore, coding "Fair" as non-support provides a conservative test of committed incumbent support, distinguishing explicit support from ambivalent and biased responses.

The independent variable of interest is exposure to victimization, operationalized in two distinct forms to capture the dual sources of violence in El Salvador:

\noindent Independent Variable 1 ($X_{Gang}$): Measured as exposure to non-state extortion (Gang Extortion) using variable $VICBAR4A$. This variable captures direct victimization by criminal groups. The operationalization is as follows:

\begin{spacing}{1}
\begin{equation}
\text{Gang Victim} =
\begin{cases}
1, & \text{if respondent/family was a victim of extortion in last 12 months} \\
0, & \text{otherwise}
\end{cases}
\end{equation}
\end{spacing}

\vspace{1em}

\noindent Independent Variable 2 ($X_{State}$): Measures institutional abuse as exposure to state-led "kickback" (Bribery) using variable EXC6. This variable captures direct institutional corruption:

\begin{spacing}{1}
\begin{equation}
\text{Bribe Victim} =
\begin{cases}
1, & \text{if a public official requested a bribe in last 12 months} \\
0, & \text{otherwise}
\end{cases}
\end{equation}
\end{spacing}

\vspace{1em}

\noindent Moderator ($Z$): Economic Perception - Measured as the respondent's retrospective evaluation of their personal economic situation (IDIO2), serving as a proxy for the receipt of effective state investment. Coded on a 3-point ordinal scale:

\begin{spacing}{1}
\begin{equation}
\text{Econ Perception} =
\begin{cases}
1, & \text{Worse} \\
2, & \text{Same} \\
3, & \text{Better}
\end{cases}
\end{equation}
\end{spacing}

\vspace{1em}

\noindent Control Variables (C): To mitigate confounding, the following structural and political variables are included:

\noindent Education ($Inc_{proxy}$): A proxy for Socio-Economic Status (SES). Coded as a 4-point ordinal scale based on the highest level of education completed (edr):

\begin{spacing}{1}
\begin{equation}
\text{Education Level} =
\begin{cases}
0, & \text{None}, \\
1, & \text{Primary}, \\
2, & \text{Secondary}, \\
3, & \text{Higher Education}
\\
\end{cases}
\end{equation}
\end{spacing}

\vspace{1em}

\noindent Urbanization ($U$): Binary indicator for urban residence (ur1new), accounting for the concentration of both extortion activity and economic opportunity.

\begin{spacing}{1}
\begin{equation}
\text{Urban} =
\begin{cases}
1, & \text{if residence is Urban (City/Outskirts)} \\
0, & \text{if residence is Rural}
\\
\end{cases}
\end{equation}
\end{spacing}

\vspace{1em}

\noindent Corruption Perception ($C$): A generalized measure of how common the respondent believes corruption is among politicians (exc7). Reversed so higher values indicate higher cynicism:

\begin{spacing}{1}
\begin{equation}
\text{High Corruption Perc.} =
\begin{cases}
1, & \text{Very Uncommon} \\
2, & \text{Uncommon} \\
3, & \text{Common} \\
4, & \text{Very Common}
\end{cases}
\end{equation}
\end{spacing}

\vspace{1em}

\noindent Municipal Services ($W_{proxy}$): Satisfaction with local services (sgl1), serving as a proxy for the tangible delivery of public goods by the national government. Inverted so higher values indicate better services:

\begin{spacing}{1}
\begin{equation}
\text{Muni Services} =
\begin{cases}
1, & \text{Very Bad} \\
2, & \text{Bad} \\
3, & \text{Fair} \\
4, & \text{Good} \\
5, & \text{Very Good}
\end{cases}
\end{equation}
\end{spacing}

## Results

Results will be written up soon. In short I find that because the data is from 2021, it is hard to say much about democratic backsliding / transition (since the data is after the 2019 election of Bukele). That said I find those who are exposed to gang violence blame Bukele suggesting that voters point to the current ruling party despite their claims of success.

### Model Estimation

```{r}
#| output: asis
#| message: false

#Run logistic regression for H1
model_base <- glm(bukele_support ~ 
                    extortion_victim +             # IV
                    econ_perception +              # Moderator 
                    is_urban + 
                    edr + 
                    corruption_perc + 
                    muni_services,
                  data = LAPOP_Model, 
                  family = "binomial")

# 2. Run the Interaction Model (Testing H1)
# This adds the "Dampening" test
model_interaction <- glm(bukele_support ~ 
                           extortion_victim + 
                           econ_perception +
                           is_urban + 
                           edr + 
                           corruption_perc + 
                           muni_services +
                           extortion_victim * econ_perception ,
                         data = LAPOP_Model, 
                         family = "binomial")

# 3. Create the Side-by-Side Table
stargazer(model_base, model_interaction, 
          type = "latex",
          title = "Logistic Regression Results: Determinants of Bukele Support",
          column.labels = c("Main Effects", "Interaction (H3)"),
          
          # Labels for your variables
          covariate.labels = c("Extortion Victim", 
                               "Econ Perception (Better)", 
                               "Urban (Yes)", 
                               "Education (0-3)", 
                               "Corruption Perc (High)", 
                               "Muni Services (Better)",
                               "Interaction: Victim x Econ"),
          
          dep.var.labels = "Support for Bukele",
          star.cutoffs = c(0.05, 0.01, 0.001))

```

```{r}
#| output: asis

# HA3
model_bribe_base <- glm(bukele_support ~ 
                          bribe_victim + 
                          econ_perception + 
                          is_urban + 
                          edr + 
                          corruption_perc + 
                          muni_services,
                          data = LAPOP_Model, family = "binomial",
                        )

model_bribe_interaction <- glm(bukele_support ~ 
                                 bribe_victim * econ_perception + 
                                 is_urban + 
                                 edr + 
                                 corruption_perc + 
                                 muni_services,
                               data = LAPOP_Model, family = "binomial")

# 3. View Results
stargazer(model_bribe_base, model_bribe_interaction, type = "latex",
          covariate.labels = c("Bribe Victim (State)", "Econ Perception (Better)", 
                               "Urban", "Education", "Gen. Corruption Perc", "Muni Services",
                               "Interaction: Bribe x Econ"),
          dep.var.labels = "Support for Bukele")
```

\newpage

# Limitations

Note: This submission focuses on the LAPOP survey data from 2021. That said, there are a couple of limitations with this data set that need to be addressed before the finalization of this paper.

Currently, I am examining different approaches to aggregation and measures of my variables using the following data sets:

2021 Legislative Election Data (n = 262)

-   Municipality Vote share for each party
-   Bukele’s Party as opposed to direct vote

DIGESTYC - Census (Similar to ACS) 2021 - Municipality level demographic data

ACLED Conflict Data - Gang and State Violence events from 2018 to 2024

CSES 2019 Post-Election Survey (El Salvador) - Comparative Study of Electoral Systems (CSES)

LAPOP 2023 Survey Data (which has questions about the 2019 election which is far more interesting and relevant but seems to be biased)

\newpage

## Works Cited
